---
title: 'Deshpande_Shishir'
output:
  html_document: default
---

```{r setup, include=FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

library(moments)  # install.packages("moments")

```

-----

### Test Items starts from here - There are 5 sections - 75 points total ##########################

##### Section 1: (15 points) ##################################

##### (1) R has probability functions available for use (Kabacoff, Section 5.2.3). Using one distribution to approximate another is not uncommon.

(1)(a) (6 points) The Poisson distribution may be used to approximate the binomial distribution if n > 20 and np < 7. Estimate the following binomial probabilities using *dpois()* or *ppois()* with probability p = 0.05, and n = 100. Then, estimate the same probabilities using *dbinom()* or *pbinom()*.  Show the numerical results of your calculations.

(i) The probability of exactly 0 successes.

```{r test1ai}
#Probability calculation
poisson <- dpois(0, 0.05*100)
binomial <- dbinom(0, 100, 0.05)

#Displaying results
poisson
binomial

```

(ii) The probability of fewer than 7 successes. Please note the following, taken from the Binomial Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aii}
# The Poisson distribution is used to approximate the binomial when: (1) n is large, (2) p is small, & (3) np is moderate

# Calculating probabilities with max 6 successes
poisson2 <- ppois(6, 0.05*100)
binomial2 <- pbinom(6, 100, 0.05)

#Displaying results
poisson2
binomial2

```

The binomial may also be approximated via the normal distribution. Estimate the following binomial probabilities using *dnorm()* or *pnorm()*, this time with probability p = 0.2 and n = 100. Then, calculate the same probabilities using *dbinom()* and *pbinom()*. Use continuity correction. Show the numerical results of your calculations.

(iii) The probability of exactly 25 successes.

```{r test1aiii}
# For normal distribution, we will need to calculate Mean (n * p) & SD (sqrt(n * p * (1 - p)))
mean <- 100 * 0.2
SD <- sqrt(100 * 0.2 * (1 - 0.2))

# Calculating probability for normal distribution
normal_prob <- pnorm(25.5, mean, SD) - pnorm(24.5, mean, SD)

# Calculating binomial probability
binomial_prob <- dbinom(25, size = 100, prob = 0.2)

# Displaying results
normal_prob
binomial_prob

```

(iv) The probability of fewer than 25 successes. Please note the following, taken from the Normal Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aiv}
# Normal probability
normal_prob_less_25 <- pnorm(24.5, mean, SD, lower.tail = TRUE) # Note: I use 24.5 to approximate probability of less than 25 successes

# Binomial calculation
binomial_prob_less_25 <- pbinom(24, size = 100, prob = 0.2) # Note: I use 24 since binomial distribution considers whole numbers

# Display results
normal_prob_less_25
binomial_prob_less_25

```

(1)(b) (3 points) Generate side-by-side barplots using *par(mfrow = c(1,2))* or *grid.arrange()*. The left barplot will show Poisson probabilties for outcomes ranging from 0 to 10. The right barplot will show binomial probabilities for outcomes ranging from 0 to 10. Use p = 0.1 and n = 100. Title each plot,  present in color and assign names to the bar; i.e. x-axis value labels.

```{r test1b}
# Side by side plot
par(mfrow = c(1, 2))

# Setting parameters
p <- 0.1
n <- 100
outcome_range <- 0:10

# Probability calculation
poisson3 <- dpois(outcome_range, 100*0.1)
binomial3 <- dbinom(outcome_range, size = n, prob = p)

# Plotting
barplot(poisson3, names.arg = outcome_range, col = "lightblue", main = "Poisson Probabilities", xlab = "Outcomes", ylab = "Probability")
barplot(binomial3, names.arg = outcome_range, col = "lightgreen", main = "Binomial Probabilities", xlab = "Outcomes", ylab = "Probability")

```

(1)(c) (6 points) For this problem, refer to Sections 5.2 of Business Statistics. A discrete random variable has outcomes:  0, 1, 2, 3, 4, 5, 6.  The corresponding probabilities in sequence with the outcomes are: 0.215, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001.  In other words, the probabilty of obtaining "0" is 0.215.  

(i)  Calculate the expected value and variance for this distribution using the general formula for mean and variance of a discrete distribution. To do this, you will need to use integer values from 0 to 6 as outcomes along with the corresponding probabilities. Round your answer to 1 decimal place. 

```{r test1ci}
# Parameters
var_out <- c(0, 1, 2, 3, 4, 5, 6)
prob_seq <- c(0.215, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001)

# Calculations
exp_value <- round(sum(var_out * prob_seq), 1)
variance_value <- round((sum(var_out^2 * prob_seq) - exp_value^2), 1)

# Displaying results
exp_value
variance_value

```

(ii) Use the *cumsum()* function and plot the cumulative probabilties versus the corresponding outcomes. Detemine the value of the median for this distribution and show on this plot. Note that there are methods for interpolating a median. However, we can identify an appropriate median from our set of our outcomes - 0 through 6 - that satisfies the definition. Creating a stair-step plot of the cumulative probability as a function of the outcomes may be helpful in identifying it.

```{r test1cii}
# Creating a cumulative sum data frame with values and cumulative probability
cp_df <- data.frame(Values = var_out, Probabilities = cumsum(prob_seq))

# Finding median value and median cumulative probability
median_val <- median(cp_df$Values)
median_prob <- median(cp_df$Probabilities)


# Plotting
plot(cp_df$Values, cp_df$Probabilities, col = "red", main = "Cumulative Probabilities vs. Outcomes", type = "s", 
     xlab = "Outcomes", ylab = "Cumulative Probability")

# Adding median point and line to plot
abline(h = median_prob, col = "green", lty = 2)
abline(v = median_val, col = "green", lty = 2)
text(median_val, median_prob - 0.1, 
     labels = paste("Median Value =", median_val, 
                    "\nMedian Probability =", median_prob), 
     pos = 4, col = "darkgreen")


```

##### Section 2: (15 points) ##################################

#####  (2) Conditional probabilities appear in many contexts and, in particular, are used by Bayes' Theorem. Correlations are another means for evaluating dependency between variables. The dataset "faithful"" is part of the "datasets" package and may be loaded with the statement *data(faithful)*. It contains 272 observations of 2 variables;  waiting time between eruptions (in minutes) and the duration of the eruption (in minutes) for the Old Faithful geyser in Yellowstone National Park.

(2)(a) (6 points) Load the "faithful" dataset and present summary statistics and a histogram of waiting times.  Additionally, compute the empirical conditional probability of an eruption less than 3.5 minutes, if the waiting time exceeds 70 minutes.

```{r test2a}
# Loading dataset
data(faithful, package = "datasets")

# Presenting summary statistics
summary(faithful)

#Histogram of waiting times
hist(faithful$waiting, col ="lightgreen", main = "Histogram of Waiting Times (minutes)", 
     xlab = "Waiting Time (Minutes)", ylab = "Frequency of Eruptions")

# Conditional probability for eruption
waiting_70 <- subset(faithful, waiting > 70)
erupting_3.5 <- subset(waiting_70, waiting_70$eruptions < 3.5)
conditional_prob <- nrow(erupting_3.5)/nrow(waiting_70)

# Results
cat("\nConditional Probability of an eruption less than 3.5 minutes, if waiting time exceeds 70 minutes = ", conditional_prob)

```

(i) Identify any observations in "faithful" for which the waiting time exceeds 90 minutes and the eruptions last longer than 5 minutes.  List and show any such observations in a distinct color on a scatterplot of all eruption (vertical axis) and waiting times (horizontal axis). Include a horizontal line at eruption = 5.0, and a vertical line at waiting time = 90.  Add a title and appropriate text. 

```{r test2ai}
# Observations with waiting time > 90 minutes and eruption > 5 minutes
subset90_5 <- subset(faithful, faithful$waiting > 90 & faithful$eruptions > 5)

# Displaying observations
subset90_5

# Plotting all eruptions and waiting times
plot(faithful$waiting, faithful$eruptions, col = "pink", 
     main = "Eruption Times vs. Waiting Times", xlab = "Waiting Time (minutes)", ylab = "Eruption Time (minutes)")

# Highlighting observations with waiting time > 90 minutes and eruption > 5 minutes
points(subset90_5$waiting, subset90_5$eruptions, col = "red", pch = 16)

# Add identifying lines and text
abline(h = 5, col = "blue", lty = 2)
abline(v = 90, col = "blue", lty = 2)
text(subset90_5$waiting[1] - 3.25, subset90_5$eruptions[1] + 0.03, 
     labels = paste("(", subset90_5$waiting[1], ",", subset90_5$eruptions[1], ")"), 
     col = "black", cex = 0.8)

```

(ii)  What does the plot suggest about the relationship between eruption time and waiting time?

***Answer: The plot suggests that a lightly positive relationship exists between eruption times and waiting times. However, the pattern seems to be non-linear, with two distinct clusters: (1) Those with shorter eruptions (around 2.5 minutes) and shorter waiting times (<70 minutes). (2) Those with longer eruptions (> 3.5 minutes) and longer waiting times (75 to 90 minutes). To me, this reflects a bi modal distribution that suggests the eruptions are either short and in quick succession, or long and have lower frequency in between. ***

-----

(2)(b) (6 points) Past research indicates that the waiting times between consecutive eruptions are not independent.  This problem will check to see if there is evidence of this. Form consecutive pairs of waiting times.  In other words, pair the first and second waiting times, pair the third and fourth waiting times, and so forth.  There are 136 resulting consecutive pairs of waiting times.  Form a data frame with the first column containing the first waiting time in a pair and the second column with the second waiting time in a pair. Plot the pairs with the second member of a pair on the vertical axis and the first member on the horizontal axis.

One way to do this is to pass the vector of waiting times - faithful$waiting - to *matrix()*, specifying 2 columns for our matrix, with values organized by row; i.e. byrow = TRUE.

```{r test2b}
# Form consecutive pairs of waiting times
waiting_pair <- matrix(faithful$waiting, ncol = 2, byrow = TRUE)

# Convert to a data frame
waiting_pair_df <- data.frame(First = waiting_pair[, 1], Second = waiting_pair[, 2])

# Plotting the pairs
plot(waiting_pair_df$First, waiting_pair_df$Second, main = "Waiting Time Pairs", 
     xlab = "First Waiting Time (minutes)", ylab = "Second Waiting Time (minutes)", col = "blue", pch = 16)

```

(2)(c) (3 points) Test the hypothesis of independence with a two-sided test at the 99% confidence level using the Kendall correlation coefficient. The *cor.test()* function can be used to structure this test and specify the appropriate - Kendall's tau - method.

```{r test2c}
# Testing using Kendall's test
kendall_test <- cor.test(waiting_pair_df$First, waiting_pair_df$Second, method = "kendall", conf.level = 0.99)

# Displaying results
kendall_test

```

##### Section 3: (15 points) ##################################

##### (3)  Performing hypothesis tests using random samples is fundamental to statistical inference. The first part of this problem involves comparing two different diets. Using "ChickWeight" data available in the base R, "datasets" package, we will create a subset of the "ChickWeight" data frame. Specifically, we want to create a data frame that includes only those rows where Time == 21 AND Diet == 1 or 3.

```{r test3}

# load "ChickWeight" dataset
data(ChickWeight, package = "datasets")

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.
result <- subset(ChickWeight, ChickWeight$Time == 21 & (ChickWeight$Diet == 1 | ChickWeight$Diet == 3))
head(result)

# The values in your subsetted data frame should match those below:
# > head(df)
#    weight Time Chick Diet
# 12    205   21     1    1
# 24    215   21     2    1
# 36    202   21     3    1
# 48    157   21     4    1
# 60    223   21     5    1
# 72    157   21     6    1

```

##### The data frame, "result", has chick weights for two diets, identified as diet "1" and "3". Use the data frame, "result," to complete the following item.

(3)(a) (3 points) Display two side-by-side vertical boxplots using par(mfrow = c(1,2)).  One boxplot would display Diet "1" and the other Diet "3". 

```{r test3a}
# Setting up plotting
par(mfrow= c(1, 2))

# Subsetting the data
diet_1 <- subset(result, Diet == 1)
diet_3 <- subset(result, Diet == 3)

# Boxplot for Diet 1
boxplot(diet_1$weight,
        main = "Diet 1",
        ylab = "Weight (grams)",
        col = "brown")

# Boxplot for Diet 3
boxplot(diet_3$weight,
        main = "Diet 3",
        ylab = "Weight (grams)",
        col = "grey")

```

(3)(b) (3 points)  Use the "weight" data for the two diets to test the null hypothesis of equal population mean weights for the two diets. Test at the 95% confidence level with a two-sided t-test. This can be done using *t.test()* in R. Do not assume equal variances. Display the results of *t.test()*.

```{r test3b}
# Testing two-sided t-test at 95% confidence
t_test <- t.test(diet_1$weight, diet_3$weight, var.equal = FALSE, conf.level = 0.95)

#Displaying the result
t_test

```

##### Working with paired data is another common statistical activity. The "ChickWeight" data will be used to illustrate how the weight gain from day 20 to 21 may be analyzed. This time, we will look only at those individuals on Diet == "3". You will need to add code below creating two (2) vectors. One (1) vector should include all the Time == 20 weights of those individuals on Diet == "3"; a second should include all the Time == 21 weights of those individuals on Diet == "3".

```{r test3paired}
# Subsetting the data
day20_weights <- ChickWeight$weight[ChickWeight$Diet == 3 & ChickWeight$Time == 20]
day21_weights <- ChickWeight$weight[ChickWeight$Diet == 3 & ChickWeight$Time == 21]

# First six elements of both vectors
cat("Vector 1 with first 6 elements for Time == 20 and Diet == 3\n")
head(day20_weights, 6)
cat("\nVector 2 with first 6 elements for Time == 21 and Diet == 3\n")
head(day21_weights, 6)

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.

# The first six (6) elements of your Time == 20 vector should match those below:
# [1] 235 291 156 327 361 225

```

(3)(c) (3 points) Present a scatterplot of the Time == 21 weights as a function of the Time == 20 weights.  Include a diagonal line with zero intercept and slope equal to one. Title and label the variables in this scatterplot.  

```{r test3c}
# Plotting Time == 21 weights as a function of Time == 20 weights
plot(day20_weights, day21_weights, 
  main = "Weight at Day 21 vs Day 20", 
  xlab = "Weight at Day 20 (grams)", 
  ylab = "Weight at Day 21 (grams)", 
  col = "blue", pch = 16)

# Add diagonal line with zero intercept and slope of 1
abline(a = 0, b = 1, col = "red", lty = 2)

```

(3)(d) (6 points) Calculate and present a one-sided, 95% confidence interval for the average weight gain from day 20 to day 21. Write the code for the paired t-test and for determination of the confidence interval endpoints. **Do not use *t.test()**, although you may check your answers using this function. Present the resulting test statistic value, critical value, p-value and confidence interval.

```{r test3d}
# Calculating weight gain
weight_gain <- day21_weights - day20_weights

# Calculating number of observations
num_obs <- length(weight_gain)

# Calculating mean and Standard Deviation
mean_gain <- mean(weight_gain)
SD <- sd(weight_gain)

# Calculating test statistic
test_stat <- mean_gain / (SD / sqrt(num_obs))

# Calculating degrees of freedom
dof <- num_obs - 1

# Critical value at 95% confidence interval
crit_val <- qt(0.95, dof, lower.tail = TRUE)

# Lower bound of confidence interval
low_conf <- mean_gain - crit_val * (SD / sqrt(num_obs))

# Calculating p-value
p_val <- pt(test_stat, dof, lower.tail = FALSE)

# Display results
list(
  mean_statistic = mean_gain,
  test_statistic = test_stat,
  critical_value = crit_val,
  degrees_of_freedom = dof,
  p_value = p_val,
  confidence_interval = c(low_conf, Inf)
)

# For comparison
t_test_result <- t.test(day21_weights, day20_weights,
                        paired = TRUE,
                        alternative = "greater",
                        conf.level = 0.95)

print("T-test for comparison")
print(t_test_result)

```

##### Section 4: (15 points) ##################################

##### (4) Statistical inference depends on using a sampling distribution for a statistic in order to make confidence statements about unknown population parameters. The Central Limit Theorem is used to justify use of the normal distribution as a sampling distribution for statistical inference. Using Nile River flow data from 1871 to 1970, this problem demonstrates sampling distribution convergence to normality. Use the code below to prepare the data.  Refer to this example when completing (4)(c) below.

```{r test4}
data(Nile, package = "datasets")

# Central statistics
num_Nile <- length(Nile)
mean_Nile <- mean(Nile)
sd_Nile <- sd(Nile)

# Examining the distribution visually
x <- Nile
hist(Nile, freq = FALSE, col = "lightblue", xlab = "Flow",
     main = "Histogram of Nile River Flows, 1871 to 1970")
curve(dnorm(x, mean = mean_Nile, sd = sd_Nile), col = "black", lwd = 2, add = TRUE)

# The distribution shows a slight right skew, suggesting the data is somewhat normally distributed. We can confirm by calculating skewness and the kurtosis values

skewness_value <- skewness(Nile)
kurtosis_value <- kurtosis(Nile) - 3 # Subtracting 3 to account for excess kurtosis

# Display skewness and kurtosis
skewness_value
kurtosis_value

# The skewness being slightly higher than 0 and kurtosis value being slightly lower than 0 confirms that the data is somewhat normally distributed. Therefore, it is reasonable to assume normality when applying parametric statistical methods such as confidence intervals 

```

(4)(a) (3 points) Using Nile River flow data and the "moments" package, calculate skewness and kurtosis. Present a QQ plot and boxplot of the flow data side-by-side using *qqnorm()*, *qqline()* and *boxplot()*; *par(mfrow = c(1, 2))* may be used to locate the plots side-by-side. Add features to these displays as you choose.

```{r test4a}
# Loading library
library(moments)

# Calculating skewness & kurtosis
skewness(Nile)
kurtosis(Nile)

# Setting up plot
par(mfrow = c(1, 2))

# QQ plots
qqnorm(Nile, main = "QQ plot of Nile Data", col = "lightblue", pch = 16)
qqline(Nile, col = "black", lwd = 2)

# Box plots
boxplot(Nile, main = "Boxplot of Nile data", col = "lightgreen", ylab = "Nile flow")

```

(4)(b) (6 points) Using *set.seed(456)* and the Nile data, generate 1000 random samples of size n = 16, with replacement. For each sample drawn, calculate and store the sample mean. This can be done with a for-loop and use of the *sample()* function. Label the resulting 1000 mean values as "sample1". **Repeat these steps using *set.seed(789)* - a different "seed" - and samples of size n = 64.** Label these 1000 mean values as "sample2". Compute and present the means, sample standard deviations and sample variances for "sample1" and "sample2" in a table with the first row for "sample1", the second row for "sample2" and the columns labled for each statistic.

```{r test4b}
# Set 1
set.seed(456)
sample1 <- numeric(1000)
# Looping
for(i in 1:1000) {
  sample1[i] <- mean(sample(Nile, size = 16, replace = TRUE))
}

# Set 2
set.seed(789)
sample2 <- numeric(1000)
# Looping
for(i in 1:1000) {
  sample2[i] <- mean(sample(Nile, size = 64, replace = TRUE))
}

# Statistics for sample1 & sample2
mean_sample1 <- mean(sample1)
sd_sample1 <- sd(sample1)
var_sample1 <- var(sample1)

mean_sample2 <- mean(sample2)
sd_sample2 <- sd(sample2)
var_sample2 <- var(sample2)

# Tabulating results
results <- data.frame(
  Statistic = c("Mean", "Standard Deviation", "Variance"),
  Sample1 = c(mean_sample1, sd_sample1, var_sample1),
  Sample2 = c(mean_sample2, sd_sample2, var_sample2)
)

# Displaying results
results


```

(4)(c) (6 points) Present side-by-side histograms of "sample1" and "sample2" with the normal density curve superimposed. To prepare comparable histograms, it will be necessary to use "freq = FALSE" and to maintain the same x-axis with "xlim = c(750, 1050)", and the same y-axis with "ylim = c(0, 0.025)." **To superimpose separate density functions, you will need to use the mean and standard deviation for each "sample" - each histogram - separately.** 

```{r test4c}

# Create histograms of "sample1" and "sample2" with normal density curves superimposed
par(mfrow = c(1, 2))
hist(sample1, freq = FALSE, xlim = c(750, 1050), ylim = c(0, 0.025), col = "lightblue")
curve(dnorm(x, mean = mean(sample1), sd = sd(sample1)), col = "black", lwd = 2, add = TRUE)
hist(sample2, freq = FALSE, xlim = c(750, 1050), ylim = c(0, 0.025), col = "lightblue")
curve(dnorm(x, mean = mean(sample2), sd = sd(sample2)), col = "black", lwd = 2, add = TRUE)

```

-----

##### Section 5: (15 points) ##################################

##### (5)  This problem deals with contingency table analysis. This is an example of categorical data analysis (see Kabacoff, pp. 145-151). The "warpbreaks" dataset gives the number of warp breaks per loom, where a loom corresponds to a fixed length of yarn.  There are 54 observations on 3 variables: breaks	(numeric, the number of breaks), wool (factor, type of wool: A or B), and tension (factor, low L, medium M and high H).  These data have been studied and used for example elsewhere.  For the purposes of this problem, we will focus on the relationship between breaks and wool using contingency table analysis.

(5)(a)(5 points) warpbreaks is part of the "datasets" package and may be loaded via *data(warpbreaks)*.  Load "warpbreaks" and present the structure using *str()*. Calculate the median number of breaks for the entire dataset, disregarding "tension" and "wool". Define this median value as "median_breaks". Present a histogram of the number of breaks with the location of the median indicated.

Create a new variable "number" as follows:  for each value of "breaks", classify the number of breaks as either strictly below "median_breaks", or the alternative. Convert the "above"|"below" classifications to a factor, and combine with the dataset warpbreaks.  Present a summary of the augmented dataset using *summary()*.  Present a contingency table of the frequency of breaks using the two variables "wool" and "number".  There should be four cells in this table.

```{r test5a}
data(warpbreaks, package = "datasets")

# Structure
str(warpbreaks)

# Calculating median
median_breaks <- median(warpbreaks$breaks)
median_breaks

# Histogram with median
hist(warpbreaks$breaks, col = "lightblue", main = "Histogram of breaks",
     xlab = "Number of Breaks", ylab = "Frequency")
abline(v = median_breaks, col = "black", lwd = 2, lty = 2)
text(24, 15,
     labels = paste("Median", median_breaks), 
     col = "black", cex = 0.8, srt = 90)

# Creating new variable "number"
warpbreaks$number <- ifelse(warpbreaks$breaks < median_breaks, "below", "above")
warpbreaks$number <- factor(warpbreaks$number, levels = c("above", "below"))

# Summary of augmented dataset
summary(warpbreaks)

# Contingency table
conting_table <- table(warpbreaks$wool, warpbreaks$number)
conting_table

```

(5)(b)(3 points)  Using the table constructed in (5)(a), test at the 5% level the null hypothesis of independence using the uncorrected *chisq.test()* (Black, Business Statistics, Section 16.2). Show the results of this test and state your conclusions. 

```{r test5b}
chisq.test(conting_table, correct=FALSE)

# Post run comment: The chi-squared test shows that the X-squared value for the contingency table is larger than its p-value. Through this, we can conclude that the variables are not independent of each other and reject the null hypothesis.

```


(5)(c) (3 points) 'Manually' calculate the chi-squared statistic and p-value of the table from (5)(a). The *addmargins()* function can be used to add row and column sums to the table; useful for calculating the expected values for each cell. You should be able to match the chi-squared and p-values from (5)(b). The underlying code for the *chisq.test()* function can be viewed by entering *chisq.test* - without parentheses - in the Console. You are given code below to create the table, add row and column sums and calculate the expected values for the for the first of two (2) rows. You will need to add code to calculate the expected values for the second row and the chi-squared. The *pchisq()* function can be used to return the p-value.

```{r test5c}

tbl <- table(warpbreaks$wool, warpbreaks$number)
mar_tbl <- addmargins(tbl)

e11 <- mar_tbl[3, 1] * mar_tbl[1, 3] / mar_tbl[3, 3]
e12 <- mar_tbl[3, 2] * mar_tbl[1, 3] / mar_tbl[3, 3]

# Expected values for 2nd row
e21 <- mar_tbl[3, 1] * mar_tbl[2, 3] / mar_tbl[3, 3]
e22 <- mar_tbl[3, 2] * mar_tbl[2, 3] / mar_tbl[3, 3]

# Putting expected values into a matrix
expected_values <- matrix(c(e11, e12), nrow = 2, ncol = 2, byrow = TRUE)

# Chi-squared statistic
chi_squared <- sum((conting_table - expected_values)^2 / expected_values)

# Degrees of freedom
dof <- (nrow(tbl) - 1)^2

# Calculate the p-value
p_value <- pchisq(chi_squared, df = dof, lower.tail = FALSE)

# Display results
list(chi_squared = chi_squared, p_value = p_value, degrees_of_freedom = dof)

```

(5)(d) (4 points) Build a user-defined function, using your code for (5)(c).We want to pass our (5)(a) table to our function and have it return the chi-squared statistic and p-value. You're provided with the 'shell' of a function and will need to add code to calculate the expected values, the chi-squared statistic, the p-value and return (i.e. output) the chi-squared and p-value.

```{r 5d}

chisq_function <- function(x) {
  # Code for calculating the expected values
  mar_tbl <- addmargins(x)
  e11 <- mar_tbl[3, 1] * mar_tbl[1, 3] / mar_tbl[3, 3]
  e12 <- mar_tbl[3, 2] * mar_tbl[1, 3] / mar_tbl[3, 3]
  e21 <- mar_tbl[3, 1] * mar_tbl[2, 3] / mar_tbl[3, 3]
  e22 <- mar_tbl[3, 2] * mar_tbl[2, 3] / mar_tbl[3, 3]
  
  # Code for calculating the chi-squared
  chi_squared <- sum((x[1, 1] - e11)^2 / e11,
                     (x[1, 2] - e12)^2 / e12,
                     (x[2, 1] - e21)^2 / e21,
                     (x[2, 2] - e22)^2 / e22)
  
  # Code for calculating the degrees of freedom and p-value
  dof <- (nrow(x) - 1)^2
  p_value <- pchisq(chi_squared, df = dof, lower.tail = FALSE)
    
  # Code to output the chi-squared, degrees of freedom and p-value 
  return(list(chi_squared = chi_squared,
              degrees_of_freedom = dof,
              p_value = p_value))
  
}

chisq_function(tbl)

```

You do not need to do anything with the below. It is provided only for demonstration purposes. In (5)(d), we know the size of the table - 2 x 2 - and write a function to match. Often, though, we'll want to write functions that are flexible in some way.

```{r chisq_vectorized}
# Below is a function that should return the same values as chisq.test() and your
# function from (5)(d). Here, though, the function loops over the rows and columns
# to calculate the expected values. Ideally, this function would work for any sized
# table.

chisqfun <- function(t) {
   x <- addmargins(t)
   e <- matrix(0, nrow = nrow(t), ncol = ncol(t), byrow = T)
   r <- matrix(0, nrow = nrow(t), ncol = ncol(t), byrow = T)
   for (i in 1:dim(t)[1]) {
       for (j in 1:dim(t)[2]) {
          e[i,j] = x[nrow(x),j] * x[i,ncol(x)]/x[nrow(x), ncol(x)]
         r[i,j] = ((x[i,j] - e[i,j])^2)/e[i,j]
         }
     }
  chi <- sum(r)
  xdf <- (nrow(t) - 1) * (ncol(t) - 1)
  pv <- pchisq(chi, df = xdf, lower.tail = FALSE)
  return(list("chi-squared" = chi, "degrees_of_freedom" = xdf, "p-value" = pv))
  }

```
