---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Deshpande, Shishir"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (Enter your answer here.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
sex_class_table <- table(mydata$SEX, mydata$CLASS)
margin_table <- addmargins(sex_class_table)

print(margin_table)

# Barplot of data
barplot(sex_class_table, beside = TRUE, col = c("lightblue", "lightgreen", "pink"),
        legend = rownames(sex_class_table), 
        main = "Barplot of CLASS, SEX-differentiated",
        xlab = "CLASS", ylab = "Count")



```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (Enter your answer here.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
sample_data <- sample(seq(1:1036), 200)
work <- mydata[sample_data, ]

# Plotting a scatterplot for variables 2-6
plot(work[, 2:6], cex=.5, lwd=.3)

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$VOLUME, mydata$WHOLE,
     col = as.numeric(mydata$CLASS),
     pch = 20,
     xlab = "VOLUME", ylab = "WHOLE",
     main = "Plotting WHOLE weight, in grams vs. VOLUME, in cc by CLASS")

legend("bottomright", legend = levels(mydata$CLASS),
       col = 1:length(levels(mydata$CLASS)), pch = 20)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK, 
     col = as.numeric(mydata$CLASS),
     pch = 20,
     xlab = "WHOLE", ylab = "SHUCK",
     main = "Plotting SHUCK weight in grams, vs. WHOLE weight, in grams by CLASS")

legend("bottomright", legend = levels(mydata$CLASS),
       col = 1:length(levels(mydata$CLASS)), pch = 20)

# Finding max SHUCK/WHOLE ratio
max_ratio <- max(mydata$SHUCK / mydata$WHOLE, na.rm = TRUE)

# Add a line with slope = max_ratio and intercept = 0
abline(a = 0, b = max_ratio, col = "black", lwd = 2, lty = "dashed")

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: I think the first plot is more variable than the second plot. This suggests that "SHUCK" correlates to "WHOLE" better than "WHOLE" to "VOLUME". As mentioned in the question, "SHUCK" is a part of "WHOLE", so this sort of tighter clustering is expected. On the other hand, "VOLUME" is derived from the abalone's length, height and diameter which I expect to correlate less to the "SHUCK" than something that consists of the "SHUCK" itself. Also, the correlation seems to get weaker as the whole becomes larger in both plots, suggesting that the "SHUCK" is better in earlier stages of the abalone's lifespan.***
-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
# This makes 9 plots
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))  # adding margins to enhance legibility

# Histogram for M
sexM <- mydata[mydata$SEX == "M", "RATIO"]
hist(sexM, main = "Histogram - Male RATIO", col = "pink", xlab = "RATIO", 
     breaks = 25, cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Histogram for F
sexF <- mydata[mydata$SEX == "F", "RATIO"]
hist(sexF, main = "Histogram - Female RATIO", col = "lightgreen", xlab = "RATIO",
     breaks = 25, cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Histogram for I
sexI <- mydata[mydata$SEX == "I", "RATIO"]
hist(sexI, main = "Histogram - Infant RATIO", col = "lightblue", xlab = "RATIO",
     breaks = 25, cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Boxplot for M
boxplot(sexM, main = "Boxplot - Male RATIO", col = "pink", ylab = "RATIO",
        cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Boxplot for F
boxplot(sexF, main = "Boxplot - Female RATIO", col = "lightgreen", ylab = "RATIO",
        cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Boxplot for I
boxplot(sexI, main = "Boxplot - Infant RATIO", col = "lightblue", ylab = "RATIO",
        cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)

# Q-Q Plot for M
qqnorm(sexM, main = "Q-Q Plot - Male RATIO", col = "pink",
       cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)
qqline(sexM, col = "red")

# Q-Q Plot for F
qqnorm(sexF, main = "Q-Q Plot - Female RATIO", col = "lightgreen",
       cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)
qqline(sexF, col = "red")

# Q-Q Plot for I
qqnorm(sexI, main = "Q-Q Plot - Infant RATIO", col = "lightblue",
       cex.main = 1.5, cex.lab = 1.2,cex.axis = 1.2)
qqline(sexI, col = "red")

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: From the displays, it seems that the data is right-skewed, i.e., lower ratios tend to be more frequent for each sex. All three types of plots also indicate the presence of outliers that could be skewing the data and they seem more apparent in the box plots. Considering this, I would assume that these distributions are not normally distributed. Also, the QQ plots seem to rise above the line for larger values. Sex-wise, the female distribution seems to have highest skewness, but the infant distribution seems to have the most number of outliers.***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
# Deriving the outliers
male_outliers <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x= mydata$RATIO[mydata$SEX == "M"])$out, ]
female_outliers <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x= mydata$RATIO[mydata$SEX == "F"])$out, ]
infant_outliers <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x= mydata$RATIO[mydata$SEX == "I"])$out, ]

# Printing the stats
kable(rbind(male_outliers, female_outliers, infant_outliers))

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: The results show that the ratios for infant abalones have the most outliers and they all seem to come from A1 & A2 classes exclusively. In fact, almost all outliers across sexes come from A1, A2 & A3 classes, with a single A5 class outlier. Notably, the female and infant abalones also seem to have extreme outliers (row 350 for Female & row 3 for Infant). The presence of outliers, while not an insurmountable issue, does pose an issue when trying to predict ratio based on other attributes or drawing correlations to other attributes. ***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
# Plot area
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))  # adding margins to enhance legibility

# Boxplot of Volume by Class
boxplot(VOLUME ~ CLASS, data = mydata,
        main = "Volume by Class",
        xlab = "Class", ylab = "Volume",
        col = c("lightgreen", "green", "lightblue", "blue", "pink"))

# Boxplot of Whole by Class
boxplot(WHOLE ~ CLASS, data = mydata,
        main = "Whole by Class",
        xlab = "Class", ylab = "Whole",
        col = c("lightgreen", "green", "lightblue", "blue", "pink"))

# Scatterplot of Volume and Rings
plot(mydata$RINGS, mydata$VOLUME,
     main = "Volume vs Rings",
     xlab = "Rings", ylab = "Volume")

# Scatterplot of Whole and Rings
plot(mydata$RINGS, mydata$WHOLE,
     main = "Whole vs Rings",
     xlab = "Rings", ylab = "Whole")

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: I believe these variables would not perform well as predictors of age. VOLUME and WHOLE seem to correlate with the number of rings positively but the correlation is not very strong. The scatter plots indicate a trend where older classes tend to be bigger and heftier, but there's significant volatility, especially the older the abalone gets. Similarly, the boxplots reveal that higher classes have higher VOLUME and WHOLE, but there's significant overlap across classes, i.e. a 150g abalone could be an A3, A4, or A5 class, which makes it challenging to estimate its age. This indicates that VOLUME and WHOLE can estimate a possible range at best and not predict age.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
# Data aggregation
data_aggregation <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))

# Creating matrix for VOLUME, SHUCK & RATIO
volume_matrix <- matrix(data_aggregation$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

shuck_matrix <- matrix(data_aggregation$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

ratio_matrix <- matrix(data_aggregation$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

# Printing the output
kable(round(volume_matrix, 2), caption="Volume")
kable(round(shuck_matrix, 2), caption="Shuck")
kable(round(ratio_matrix, 4), caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
# First plot: Mean RATIO vs CLASS, by SEX
interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX,
                 response = mydata$RATIO,
                 xlab = "CLASS", ylab = "Mean RATIO",
                 trace.label = "SEX", main = "Mean RATIO vs CLASS")

# Second plot: Mean VOLUME vs CLASS, by SEX
interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX,
                 response = mydata$VOLUME,
                 xlab = "CLASS", ylab = "Mean VOLUME",
                 trace.label = "SEX", main = "Mean VOLUME vs CLASS")

# Third plot: Mean SHUCK vs CLASS, by SEX
interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX,
                 response = mydata$SHUCK,
                 xlab = "CLASS", ylab = "Mean SHUCK",
                 trace.label = "SEX", main = "Mean SHUCK vs CLASS")

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: These plots show RATIO,  VOLUME, and SHUCK across different classes for each sex. Looking at the plots, I have several questions. Why do SHUCK and VOLUME increase from CLASS A1 through A4 for each sex, then taper off at A5? Does this mean that CLASS A4 represents the peak abalone size before they start withering due to old age? Also, why do female abalones have higher mean SHUCK and VOLUME than males, unlike most species where the males are generally larger and heavier? Is this because they have differences in their internal organs? Further, why does RATIO peak in A1 & A2 classes and decline in all other classes for all sexes? ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
# Filtering data set to < 16 rings and subsetting into infant and adult data sets
infant <- subset(mydata, RINGS < 16 & SEX == "I")
adult  <- subset(mydata, RINGS < 16 & (SEX == "M" | SEX == "F"))

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1)) # adding margins for legibility

# Infant Volume vs Rings
boxplot(VOLUME ~ RINGS, data = infant,
        col = "pink", border = "black",
        main = "Infant Volume by Rings",
        ylab = "Volume", xlab = "Rings",
        ylim = c(0, 1100), outline = TRUE)

# Adult Volume vs Rings
boxplot(VOLUME ~ RINGS, data = adult,
        col = "lightgreen", border = "black",
        main = "Adult Volume by Rings",
        ylab = "Volume", xlab = "Rings",
        ylim = c(0, 1100), outline = TRUE)

# Infant Whole weight vs Rings
boxplot(WHOLE ~ RINGS, data = infant,
        col = "lightblue", border = "black",
        main = "Infant Whole by Rings",
        ylab = "Whole", xlab = "Rings",
        ylim = c(0, 400), outline = TRUE)

# Adult Whole weight vs Rings
boxplot(WHOLE ~ RINGS, data = adult,
        col = "orange", border = "black",
        main = "Adult Whole by Rings",
        ylab = "Whole", xlab = "Rings",
        ylim = c(0, 400), outline = TRUE)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays suggest that abalone growth increases steadily until Ring 11, with both Volume and Whole rising. This indicates that Ring 11 might be the stage at which they become physically mature. For infants, the plots show that they begin to grow from RING 5 and continue to grow until Ring 11-12. However, the plots indicate that variability increases as the abalones age, and this is especially true for adults which reach higher maximum VOLUME and WHOLE. Adults peak at higher values, suggesting these abalones continue to grow beyond infancy. But, most surprisingly, some infants have the same VOLUME and WHOLE as adults at equivalent RING levels. While this could be due to several factors, I would guess it has something to do with genetics or better access to food. Another thing could be is the possibility that some adults being labeled as infants or vice versa. ***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: There are some plausible statistical reasons as to why the study failed. I believe it is due to an overly simple approach for predicting age and classifying life stages based only on physical measurements like number of rings or volume. While these features do show a relationship to the abalone's age, particularly in their early life, they become less reliable as the abalones become older. For example, young abalones with lower weight and volume tend to have fewer rings, indicating that there is some correlation between size and age in early development. But this relationship does not consider sex as a factor, when our boxplots represent that relationship very well (see 5b, where the plots showed that females tend to start heavier and larger than males and other infants). The data does not classify infants by sex and this could be one reason why infant abalones have significant variability in their early stages that physical measurements alone can account for. The age-predictive potential of physical measurements declines further as the abalones go into CLASSES A3 through A5. The plots in 5c show that weight and volume stop correlating with the number of rings, suggesting that abalones reach a plateau in their physical growth at some point. This further supports the hypothesis that size and age are not consistently predicting their age and using physical measurements as the sole predictors becomes highly unreliable. To conclude, the study’s inconsistencies in accounting for categorical factors like infant sex or reproductive maturity, likely contributed to its failure. Predicting their age accurately might require more features and a biologically grounded framework , rather than trying to infer age from size alone.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: If I was presented with only an overall histogram and summary statistics from a sample, and no other context, I would ask myself and others several questions before accepting them as representative of the broader population. (1) How was the sample collected? Was it random or convenience-based? (2) What is the sample size as a percentage of the population? (3) What is the hypothesis being tested and why are we testing it? Why do we believe this sample is an appropriate sample? (4) What is the shape of the distribution i.e. is it normal or skewed? (5) What do the measures of central tendency look like (mean, median, mode etc.) for something that meets the criteria we are testing? (6) Were there any modifications or changes to the data i.e. was it cleaned or trimmed? (7) Was the data collected in a single period or over a period of time? what period of time is usually required to normalize seasonality, trends or outlier events? ***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: In my experience, analyzing data derived from observational studies presents several challenges, primarily due to the lack of a  controlled environment where observations may not be random or may have bias. This makes it difficult to isolate the impact of a specific variable or even variables influencing each other. As a result, causality is generally not definitive in observational studies. The best we can hope for is to find correlations within the data and explore causality as a secondary step. However, this is not to say that observational studies are useless. They are helpful in identifying patterns and generating a hypothesis to be studied and tested further. These hypotheses can be expanded into a study over longer periods of time or extrapolation to a larger population in low-information scenarios. For example, the impact of wearing a mask in stopping the spread of COVID-19 was untested in the strictest sense. However, observational studies showed that certain groups who wore masks more often spread COVID-19 less and the downside of mandating masks was low for most governments. ***
