---
title: "Assignment 7: Logistic Regression Basics"
author: "Shishir Deshpande"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readxl)
library(ggplot2)
library(dplyr)
library(knitr)

```

```{r}
# Load ICU data
icu <- read_excel("icu.xlsx")
```

```{r}
# ------------------------------------------------------------
# 1a. Codebook — built directly from the data dictionary
# ------------------------------------------------------------

codebook <- data.frame(
  Number = 1:21,
  Variable = c("ID", "STA", "AGE", "SEX", "RACE", "SER", "CAN", "CRN",
               "INF", "CPR", "SYS", "HRA", "PRE", "TYP", "FRA", "PO2",
               "PH", "PCO", "BIC", "CRE", "LOC"),
  Description = c(
    "Identification Code",
    "Vital Status",
    "Age",
    "Sex",
    "Race",
    "Service at ICU Admission",
    "Cancer Part of Present Problem",
    "History of Chronic Renal Failure",
    "Infection Probable at ICU Admission",
    "CPR Prior to ICU Admission",
    "Systolic Blood Pressure at ICU Admission",
    "Heart Rate at ICU Admission",
    "Previous Admission to an ICU Within 6 Months",
    "Type of Admission",
    "Long Bone, Multiple, Neck, Single Area, or Hip Fracture",
    "PO2 from Initial Blood Gases",
    "PH from Initial Blood Gases",
    "PCO2 from Initial Blood Gases",
    "Bicarbonate from Initial Blood Gases",
    "Creatinine from Initial Blood Gases",
    "Level of Consciousness at ICU Admission"
  ),
  Codes = c(
    "ID Number",
    "0 = Lived, 1 = Died",
    "Years",
    "0 = Male, 1 = Female",
    "1 = White, 2 = Black, 3 = Other",
    "0 = Medical, 1 = Surgical",
    "0 = No, 1 = Yes",
    "0 = No, 1 = Yes",
    "0 = No, 1 = Yes",
    "0 = No, 1 = Yes",
    "mm Hg",
    "Beats/min",
    "0 = No, 1 = Yes",
    "0 = Elective, 1 = Emergency",
    "0 = No, 1 = Yes",
    "0 = >60, 1 = <=60",
    "0 = >=7.25, 1 = <7.25",
    "0 = <=45, 1 = >45",
    "0 = >=18, 1 = <18",
    "0 = <=2, 1 = >2",
    "0 = No Coma or Deep Stupor, 1 = Deep Stupor, 2 = Coma"
  ),
  stringsAsFactors = FALSE
)

kable(codebook, caption = "ICU Study Variable Codebook (Source: ICU Data Dictionary)")

```

```{r}
# Verify all variable names in data match codebook
cat("Variables in dataset:", paste(names(icu), collapse = ", "), "\n")
cat("Variables in codebook:", paste(codebook$Variable, collapse = ", "), "\n")
cat("Any mismatches:", paste(setdiff(names(icu), codebook$Variable), collapse = ", "), "\n")
```

```{r load-data}
# Load ICU data
icu <- read_excel("icu.xlsx")

# Display structure
str(icu)
kable(summary(icu), caption = "Summary Statistics for ICU Dataset")

```
```{r}
# --- Response Variable Distribution ---
sta_freq <- as.data.frame(table(icu$STA))
colnames(sta_freq) <- c("Status (0=Lived, 1=Died)", "Frequency")
sta_freq$Proportion <- round(sta_freq$Frequency / sum(sta_freq$Frequency), 3)
kable(sta_freq, caption = "Distribution of Response Variable: STA")
```

```{r correlation-plot}
# Select only numeric columns
numeric_vars <- icu[, sapply(icu, is.numeric)]

# Correlation with STA
sta_cor <- cor(numeric_vars, numeric_vars$STA, use = "complete.obs")

# Barplot of correlations (excluding STA itself)
barplot(sta_cor[rownames(sta_cor) != "STA", 1], las = 2, col = heat.colors(nrow(sta_cor)-1))

```

```{r}
# --- Age Summary by Status ---
age_by_sta <- do.call(data.frame, aggregate(AGE ~ STA, data = icu, FUN = function(x) {
  c(Min    = round(min(x), 3),
    Q1     = round(quantile(x, 0.25), 3),
    Median = round(median(x), 3),
    Mean   = round(mean(x), 3),
    Q3     = round(quantile(x, 0.75), 3),
    Max    = round(max(x), 3))
}))
kable(age_by_sta, caption = "Age Summary by Survival Status")

# Survival rates by SEX and TYP
kable(round(prop.table(table(SEX = icu$SEX, STA = icu$STA), margin = 1), 3),
      caption = "Row Proportions: SEX (0=Male, 1=Female) vs STA")

kable(round(prop.table(table(TYP = icu$TYP, STA = icu$STA), margin = 1), 3),
      caption = "Row Proportions: TYP (0=Elective, 1=Emergency) vs STA")

```

# 1c. Visualizations
```{r}
# 1. Age distribution with density overlay
hist(icu$AGE,
     main   = "Figure 1.1: Distribution of Patient Age",
     xlab   = "Age (years)",
     col    = "lightblue",
     border = "white",
     freq   = FALSE)
lines(density(icu$AGE), col = "navy", lwd = 2)

# 2. Response variable — class imbalance check
barplot(table(icu$STA),
        main      = "Figure 1.2: Distribution of Survival Status",
        names.arg = c("Lived (0)", "Died (1)"),
        col       = c("steelblue", "tomato"),
        ylab      = "Count",
        border    = "white")

# 3. Age by survival status
boxplot(AGE ~ STA, data = icu,
        main   = "Figure 1.3: Age by Survival Status",
        xlab   = "Status",
        ylab   = "Age (years)",
        col    = c("steelblue", "tomato"),
        border = "gray30",
        names  = c("Lived (0)", "Died (1)"))

# 4. SEX vs STA
mosaicplot(table(SEX = icu$SEX, STA = icu$STA),
           main  = "Figure 1.4: Sex vs Survival Status",
           color = c("steelblue", "tomato"),
           xlab  = "Sex (0=Male, 1=Female)",
           ylab  = "Status (0=Lived, 1=Died)")

# 5. TYP vs STA
mosaicplot(table(TYP = icu$TYP, STA = icu$STA),
           main  = "Figure 1.5: Admission Type vs Survival Status",
           color = c("steelblue", "tomato"),
           xlab  = "Admission Type (0=Elective, 1=Emergency)",
           ylab  = "Status (0=Lived, 1=Died)")


```

# Task 2: SEX vs STA — 2x2 Contingency Table
## 2a. Contingency Table
```{r}
# Apply labels directly for clarity
icu$SEX_label <- factor(icu$SEX, levels = c(0, 1), labels = c("Male", "Female"))
icu$STA_label <- factor(icu$STA, levels = c(0, 1), labels = c("Lived", "Died"))

cont_sex <- table(SEX = icu$SEX_label, STA = icu$STA_label)

kable(addmargins(cont_sex),
      caption = "Table 2.1: 2x2 Contingency Table — SEX vs STA")


```

## 2b. Probabilities and Odds of Survival by Sex
```{r}
# Extract cell counts
male_lived   <- cont_sex["Male",   "Lived"]
male_died    <- cont_sex["Male",   "Died"]
female_lived <- cont_sex["Female", "Lived"]
female_died  <- cont_sex["Female", "Died"]

# Probabilities of survival (not death)
prob_male_survive   <- round(male_lived   / (male_lived   + male_died),   3)
prob_female_survive <- round(female_lived / (female_lived + female_died), 3)

# Odds of survival
odds_male   <- round(male_lived   / male_died,   3)
odds_female <- round(female_lived / female_died, 3)

# Odds ratio: Males vs Females
odds_ratio <- round(odds_male / odds_female, 3)

# Summary table
results_sex <- data.frame(
  Group             = c("Male", "Female"),
  Lived             = c(male_lived,           female_lived),
  Died              = c(male_died,            female_died),
  Total             = c(male_lived + male_died, female_lived + female_died),
  P_Survival        = c(prob_male_survive,    prob_female_survive),
  Odds_Survival     = c(odds_male,            odds_female)
)

kable(results_sex,
      col.names = c("Group", "Lived", "Died", "Total",
                    "P(Survival)", "Odds of Survival"),
      caption   = "Table 2.2: Survival Probabilities and Odds by Sex")

# Odds ratio
cat("Odds Ratio of Survival (Male vs Female):", odds_ratio, "\n")
```

## 2c. Visualizations
```{r}
# 1. Grouped barplot — counts by sex and status
barplot(cont_sex,
        beside     = TRUE,
        col        = c("steelblue", "tomato"),
        legend.text = c("Male", "Female"),
        args.legend = list(x = "topright", bty = "n"),
        main       = "Figure 2.1: Survival Status Counts by Sex",
        xlab       = "Survival Status",
        ylab       = "Count",
        names.arg  = c("Lived", "Died"),
        border     = "white")

# 2. Grouped barplot — proportions (row-normalized)
prop_sex <- prop.table(cont_sex, margin = 1)

barplot(t(prop_sex),
        beside     = TRUE,
        col        = c("steelblue", "tomato"),
        legend.text = c("Lived", "Died"),
        args.legend = list(x = "topright", bty = "n"),
        main       = "Figure 2.2: Survival Proportions by Sex",
        xlab       = "Sex",
        ylab       = "Proportion",
        names.arg  = c("Male", "Female"),
        border     = "white")

# 3. Mosaic plot — visual representation of the 2x2 table
mosaicplot(cont_sex,
           main  = "Figure 2.3: Mosaic Plot — Sex vs Survival Status",
           color = c("steelblue", "tomato"),
           xlab  = "Sex",
           ylab  = "Survival Status")

# 4. Odds ratio plot with reference line
groups    <- c("Male", "Female")
odds_vals <- c(odds_male, odds_female)

barplot(odds_vals,
        names.arg = groups,
        col       = c("steelblue", "tomato"),
        main      = "Figure 2.4: Odds of Survival by Sex",
        ylab      = "Odds of Survival",
        border    = "white",
        ylim      = c(0, max(odds_vals) * 1.2))
abline(h = 1, lty = 2, col = "gray40")  # reference: odds = 1 means 50/50
text(x = 0.7, y = odds_male   + 0.1, labels = odds_male,   cex = 0.9)
text(x = 1.9, y = odds_female + 0.1, labels = odds_female, cex = 0.9)


```

# Task 3: TYP vs STA — 2x2 Contingency Table
## 3a. Contingency Table
```{r}
# Apply labels directly for clarity
icu$TYP_label <- factor(icu$TYP, levels = c(0, 1), labels = c("Elective", "Emergency"))

cont_typ <- table(TYP = icu$TYP_label, STA = icu$STA_label)

kable(addmargins(cont_typ),
      caption = "Table 3.1: 2x2 Contingency Table — TYP vs STA")

```

## 3b. Probabilities and Odds of Survival by Admission Type
```{r}
# Extract cell counts
elective_lived  <- cont_typ["Elective",  "Lived"]
elective_died   <- cont_typ["Elective",  "Died"]
emergency_lived <- cont_typ["Emergency", "Lived"]
emergency_died  <- cont_typ["Emergency", "Died"]

# Probabilities of survival
prob_elective_survive  <- round(elective_lived  / (elective_lived  + elective_died),  3)
prob_emergency_survive <- round(emergency_lived / (emergency_lived + emergency_died), 3)

# Odds of survival
odds_elective  <- round(elective_lived  / elective_died,  3)
odds_emergency <- round(emergency_lived / emergency_died, 3)

# Odds ratio: Elective vs Emergency (elective as reference)
odds_ratio_typ <- round(odds_elective / odds_emergency, 3)

# Summary table
results_typ <- data.frame(
  Group         = c("Elective", "Emergency"),
  Lived         = c(elective_lived,            emergency_lived),
  Died          = c(elective_died,             emergency_died),
  Total         = c(elective_lived + elective_died, emergency_lived + emergency_died),
  P_Survival    = c(prob_elective_survive,     prob_emergency_survive),
  Odds_Survival = c(odds_elective,             odds_emergency)
)

kable(results_typ,
      col.names = c("Admission Type", "Lived", "Died", "Total",
                    "P(Survival)", "Odds of Survival"),
      caption   = "Table 3.2: Survival Probabilities and Odds by Admission Type")

cat("Odds Ratio of Survival (Elective vs Emergency):", odds_ratio_typ, "\n")


```

## 3c. Visualizations
```{r}
# 1. Grouped barplot — counts by admission type and status
barplot(cont_typ,
        beside      = TRUE,
        col         = c("steelblue", "tomato"),
        legend.text = c("Elective", "Emergency"),
        args.legend = list(x = "topright", bty = "n"),
        main        = "Figure 3.1: Survival Status Counts by Admission Type",
        xlab        = "Survival Status",
        ylab        = "Count",
        names.arg   = c("Lived", "Died"),
        border      = "white")

# 2. Grouped barplot — proportions (row-normalized)
prop_typ <- prop.table(cont_typ, margin = 1)

barplot(t(prop_typ),
        beside      = TRUE,
        col         = c("steelblue", "tomato"),
        legend.text = c("Lived", "Died"),
        args.legend = list(x = "topright", bty = "n"),
        main        = "Figure 3.2: Survival Proportions by Admission Type",
        xlab        = "Admission Type",
        ylab        = "Proportion",
        names.arg   = c("Elective", "Emergency"),
        border      = "white")

# 3. Mosaic plot
mosaicplot(cont_typ,
           main  = "Figure 3.3: Mosaic Plot — Admission Type vs Survival Status",
           color = c("steelblue", "tomato"),
           xlab  = "Admission Type",
           ylab  = "Survival Status")

# 4. Odds of survival by admission type with reference line
odds_vals_typ <- c(odds_elective, odds_emergency)

barplot(odds_vals_typ,
        names.arg = c("Elective", "Emergency"),
        col       = c("steelblue", "tomato"),
        main      = "Figure 3.4: Odds of Survival by Admission Type",
        ylab      = "Odds of Survival",
        border    = "white",
        ylim      = c(0, max(odds_vals_typ) * 1.2))
abline(h = 1, lty = 2, col = "gray40")
text(x = 0.7, y = odds_elective  + 0.5, labels = odds_elective,  cex = 0.9)
text(x = 1.9, y = odds_emergency + 0.5, labels = odds_emergency, cex = 0.9)


```

# Task 4: Age as a Determinant
## Task 4a: Logistic Regression Model Equations

# The logistic regression model for STA (Y) using AGE (X):
#
# Logistic Regression Model:
#
#         exp(β₀ + β₁ · AGE)
# π(x) = ─────────────────────
#        1 + exp(β₀ + β₁ · AGE)
#
# Logit Transformation:
#
# logit(π) = ln(π / (1 - π)) = β₀ + β₁ · AGE
#
# Where:
#   π(x)  = probability that STA = 1 (Died) given AGE = x
#   β₀    = intercept (log-odds of death when AGE = 0)
#   β₁    = change in log-odds of death for each one-year increase in AGE
#   logit = natural log of the odds of death
```{r}
# Display equations using cat() for clean output in the knitted document
cat("Logistic Regression Model:\n")
cat("π(x) = exp(β₀ + β₁·AGE) / (1 + exp(β₀ + β₁·AGE))\n\n")
cat("Logit Transformation:\n")
cat("logit(π) = ln(π / (1 - π)) = β₀ + β₁·AGE\n\n")
cat("Where:\n")
cat("  π(x) = P(STA = 1 | AGE = x) = probability of death given age\n")
cat("  β₀   = intercept\n")
cat("  β₁   = change in log-odds of death per one-year increase in AGE\n")

```

## Task 4b: Scatterplot of STA by AGE
```{r}
# Add jitter to STA values so overlapping points are visible
set.seed(42)
jittered_sta <- jitter(icu$STA, amount = 0.05)

plot(icu$AGE, jittered_sta,
     main   = "Figure 4b: Survival Status by Age",
     xlab   = "Age (years)",
     ylab   = "Status (0 = Lived, 1 = Died)",
     pch    = 19,
     col    = ifelse(icu$STA == 1, "tomato", "steelblue"),
     yaxt   = "n",
     ylim   = c(-0.15, 1.15))

# Custom y-axis with meaningful labels
axis(2, at = c(0, 1), labels = c("Lived (0)", "Died (1)"), las = 1)

# Reference lines at 0 and 1
abline(h = c(0, 1), lty = 2, col = "gray70")

# Add a smoothed trend line to show the underlying relationship
lines(lowess(icu$AGE, icu$STA, f = 0.8), col = "black", lwd = 2)

legend("topleft",
       legend = c("Lived (0)", "Died (1)", "LOWESS trend"),
       col    = c("steelblue", "tomato", "black"),
       pch    = c(19, 19, NA),
       lty    = c(NA, NA, 1),
       lwd    = c(NA, NA, 2),
       bty    = "n")

```

## Task 4c: Discretize AGE into AGE_CAT and Plot Proportions
```{r}
# Create AGE_CAT variable per assignment intervals
icu$AGE_CAT <- cut(icu$AGE,
                   breaks = c(15, 25, 35, 45, 55, 65, 75, 85, 95, Inf),
                   labels = 1:9,
                   right  = FALSE)   # intervals are [lower, upper)

# Verify the categorization
cat("AGE_CAT distribution:\n")
kable(table(AGE_CAT = icu$AGE_CAT),
      col.names = c("Age Category", "Count"),
      caption   = "Table 4c.1: Count of Patients per Age Category")

# Compute means and counts separately then merge
age_cat_means  <- aggregate(STA ~ AGE_CAT, data = icu, FUN = mean)
age_cat_counts <- aggregate(STA ~ AGE_CAT, data = icu, FUN = length)

# Rename before merge to avoid collision
names(age_cat_means)  <- c("AGE_CAT", "Proportion_Died")
names(age_cat_counts) <- c("AGE_CAT", "N")

age_cat_summary <- merge(age_cat_means, age_cat_counts, by = "AGE_CAT")
age_cat_summary$Proportion_Died <- round(age_cat_summary$Proportion_Died, 3)

# Add interval labels
interval_labels <- c("[15,24]", "[25,34]", "[35,44]", "[45,54]",
                     "[55,64]", "[65,74]", "[75,84]", "[85,94]", "[95+]")
age_cat_summary$Interval <- interval_labels[as.numeric(age_cat_summary$AGE_CAT)]

kable(age_cat_summary[, c("AGE_CAT", "Interval", "N", "Proportion_Died")],
      col.names = c("AGE_CAT", "Age Interval", "N", "Proportion Died"),
      caption   = "Table 4c.2: Proportion of Deaths by Age Category")

```

```{r fig.height=6, fig.width=8}
# Visualization
cat_nums  <- as.numeric(age_cat_summary$AGE_CAT)
props     <- age_cat_summary$Proportion_Died
ns        <- age_cat_summary$N
intervals <- age_cat_summary$Interval

# 1. Barplot — proportion died by age category
barplot(props,
        names.arg = intervals,
        col       = "steelblue",
        border    = "white",
        main      = "Figure 4c.1: Proportion of Deaths by Age Category",
        xlab      = "",
        ylab      = "Proportion Died",
        ylim      = c(0, 0.65),
        las       = 2,
        cex.names = 0.8,
        cex.main  = 0.95)
title(xlab = "Age Interval", line = 5.5)
abline(h = mean(icu$STA), lty = 2, col = "tomato", lwd = 2)
legend("topleft",
       legend = paste("Overall =", round(mean(icu$STA), 3)),
       lty = 2, col = "tomato", lwd = 2, bty = "n", cex = 0.75)

# 2. Point + line plot — trend across categories
plot(cat_nums, props,
     type   = "b",
     pch    = 19,
     col    = "steelblue",
     lwd    = 2,
     main   = "Figure 4c.2: Proportion Died vs Age Category",
     xlab   = "Age Category",
     ylab   = "Proportion Died (Mean STA)",
     xaxt   = "n",
     ylim   = c(0, 0.6))
axis(1,
     at       = cat_nums,
     labels   = intervals,
     las      = 2,
     cex.axis = 0.8)
abline(h   = mean(icu$STA), lty = 2, col = "tomato", lwd = 2)
text(x      = cat_nums,
     y      = props + 0.03,
     labels = paste0("n=", ns),
     cex    = 0.75,
     col    = "gray30")
legend("topleft",
       legend = paste("Overall mean =", round(mean(icu$STA), 3)),
       lty    = 2,
       col    = "tomato",
       lwd    = 2,
       bty    = "n")
```

## Task 4d: Fit Logistic Regression Model — STA ~ AGE
```{r}
# Fit the model
model_age <- glm(STA ~ AGE, data = icu, family = binomial)

# Coefficient Table
coef_table <- as.data.frame(summary(model_age)$coefficients)
coef_table <- round(coef_table, 3)
names(coef_table) <- c("Estimate", "Std. Error", "z value", "p-value")

kable(coef_table,
      caption = "Table 4d.1: Logistic Regression Coefficients — STA ~ AGE")

```

```{r}
# Odds Ratios with 95% Confidence Intervals
or_table <- as.data.frame(
  cbind(
    OR    = round(exp(coef(model_age)), 3),
    Lower = round(exp(confint(model_age)[, 1]), 3),
    Upper = round(exp(confint(model_age)[, 2]), 3)
  )
)

kable(or_table,
      col.names = c("Odds Ratio", "95% CI Lower", "95% CI Upper"),
      caption   = "Table 4d.2: Odds Ratios and 95% Confidence Intervals")

# Fitted equation
b0 <- round(coef(model_age)[1], 3)
b1 <- round(coef(model_age)[2], 3)

cat("Fitted Logistic Regression Model:\n")
cat(sprintf("logit(π) = %.3f + %.3f · AGE\n\n", b0, b1))
cat(sprintf("π(x) = exp(%.3f + %.3f · AGE) / (1 + exp(%.3f + %.3f · AGE))\n",
            b0, b1, b0, b1))
```

## Task 4e: Hypothesis Test Results
```{r}
# Wald Tests (from model summary)
cat("--- Wald Tests ---\n\n")
cat("Intercept:\n")
cat(sprintf("  H0: β₀ = 0  vs  Ha: β₀ ≠ 0\n"))
cat(sprintf("  z = %.3f,  p-value = %.3f\n\n",
            coef_table["(Intercept)", "z value"],
            coef_table["(Intercept)", "p-value"]))

cat("AGE Coefficient:\n")
cat(sprintf("  H0: β₁ = 0 (AGE has no effect on survival)\n"))
cat(sprintf("  Ha: β₁ ≠ 0 (AGE has a significant effect on survival)\n"))
cat(sprintf("  z = %.3f,  p-value = %.3f\n\n",
            coef_table["AGE", "z value"],
            coef_table["AGE", "p-value"]))

```

# Likelihood Ratio Test (G statistic) — to compare fitted model to null model (intercept only)
```{r}
model_null <- glm(STA ~ 1, data = icu, family = binomial)

lrt <- anova(model_null, model_age, test = "Chisq")

cat("--- Likelihood Ratio Test ---\n\n")
cat(sprintf("  Null deviance (intercept only): %.3f on %d df\n",
            model_null$deviance, model_null$df.residual))
cat(sprintf("  Residual deviance (AGE model):  %.3f on %d df\n",
            model_age$deviance, model_age$df.residual))
cat(sprintf("  G statistic (change in deviance): %.3f\n",
            model_null$deviance - model_age$deviance))
cat(sprintf("  Degrees of freedom: %d\n",
            model_null$df.residual - model_age$df.residual))
cat(sprintf("  p-value: %.3f\n\n",
            lrt$`Pr(>Chi)`[2]))

kable(lrt,
      digits  = 3,
      caption = "Table 4e.1: Likelihood Ratio Test — Null vs AGE Model")


```

## Task 4f: Model Fit Statistics — AIC, BIC, Deviance
```{r}
# Extract fit statistics
fit_stats <- data.frame(
  Statistic = c("Null Deviance", "Residual Deviance",
                "Change in Deviance (G)", "AIC", "BIC"),
  Value     = round(c(
    model_age$null.deviance,
    model_age$deviance,
    model_age$null.deviance - model_age$deviance,
    AIC(model_age),
    BIC(model_age)
  ), 3),
  df = c(
    model_age$df.null,
    model_age$df.residual,
    model_age$df.null - model_age$df.residual,
    NA, NA
  )
)

kable(fit_stats,
      col.names = c("Statistic", "Value", "df"),
      caption   = "Table 4f.1: Model Fit Statistics — STA ~ AGE",
      na        = "")

# Visualization — Null vs Fitted Deviance
deviances <- c(model_null$deviance, model_age$deviance)
dof       <- c(model_null$df.residual, model_age$df.residual)

barplot(deviances,
        names.arg = c("Null Model\n(Intercept Only)", "Fitted Model\n(AGE)"),
        col       = c("gray70", "steelblue"),
        border    = "white",
        main      = "Figure 4f.1: Deviance Comparison — Null vs Fitted Model",
        ylab      = "Deviance",
        ylim      = c(0, max(deviances) * 1.15))
text(x      = c(0.7, 1.9),
     y      = deviances * 1.03,
     labels = round(deviances, 3),
     cex    = 0.9)

```

## Task 4g: Predicted Logits
```{r}
# Predict and save logits to the dataset
icu$logit_AGE <- predict(model_age, type = "link")

# Verify logits are saved correctly
cat("First 10 predicted logits:\n")
kable(head(data.frame(ID  = icu$ID,
                      AGE = icu$AGE,
                      STA = icu$STA,
                      Logit = round(icu$logit_AGE, 3)), 10),
      caption = "Table 4g.1: Sample of Predicted Logit Values")

# Summary of predicted logits
cat("\nSummary of Predicted Logits:\n")
kable(data.frame(
  Statistic = c("Min", "Q1", "Median", "Mean", "Q3", "Max"),
  Value     = round(c(min(icu$logit_AGE),
                      quantile(icu$logit_AGE, 0.25),
                      median(icu$logit_AGE),
                      mean(icu$logit_AGE),
                      quantile(icu$logit_AGE, 0.75),
                      max(icu$logit_AGE)), 3)
),
caption = "Table 4g.2: Summary of Predicted Logit Values")

```

```{r}
# 1. Scatterplot of predicted logits by AGE — core requirement
plot(icu$AGE, icu$logit_AGE,
     main   = "Figure 4g.1: Predicted Logits by Age",
     xlab   = "Age (years)",
     ylab   = "Predicted Logit (Log-Odds of Death)",
     pch    = 19,
     col    = ifelse(icu$STA == 1, "tomato", "steelblue"),
     ylim   = c(min(icu$logit_AGE) - 0.2, max(icu$logit_AGE) + 0.2))

# Fitted line — logit is linear in AGE by definition
abline(a   = coef(model_age)[1],
       b   = coef(model_age)[2],
       col = "black",
       lwd = 2,
       lty = 1)

# Reference line at logit = 0 (i.e., π = 0.5, equal odds)
abline(h   = 0, lty = 2, col = "gray50", lwd = 1.5)

legend("topleft",
       legend = c("Lived (STA=0)", "Died (STA=1)",
                  "Fitted logit line", "logit = 0 (π = 0.50)"),
       col    = c("steelblue", "tomato", "black", "gray50"),
       pch    = c(19, 19, NA, NA),
       lty    = c(NA, NA, 1, 2),
       lwd    = c(NA, NA, 2, 1.5),
       bty    = "n")

# Annotate the age at which logit = 0 (crossover point)
age_crossover <- round(-coef(model_age)[1] / coef(model_age)[2], 1)
abline(v   = age_crossover, lty = 3, col = "gray50")
text(x      = age_crossover + 2,
     y      = min(icu$logit_AGE),
     labels = paste0("Age = ", age_crossover, "\n(π = 0.50)"),
     cex    = 0.8,
     col    = "gray30",
     adj    = 0)

# 2. Logit by AGE colored by STA with marginal rug
plot(icu$AGE, icu$logit_AGE,
     main   = "Figure 4g.2: Predicted Logits by Age with Marginal Rug",
     xlab   = "Age (years)",
     ylab   = "Predicted Logit (Log-Odds of Death)",
     pch    = 19,
     col    = ifelse(icu$STA == 1, "tomato", "steelblue"),
     ylim   = c(min(icu$logit_AGE) - 0.2, max(icu$logit_AGE) + 0.2))
abline(a   = coef(model_age)[1],
       b   = coef(model_age)[2],
       col = "black", lwd = 2)
abline(h   = 0, lty = 2, col = "gray50", lwd = 1.5)

# Rug plots show actual AGE distribution for each STA group
rug(icu$AGE[icu$STA == 0], col = "steelblue", side = 1, lwd = 1.2)
rug(icu$AGE[icu$STA == 1], col = "tomato",    side = 3, lwd = 1.2)

legend("topleft",
       legend = c("Lived (STA=0)", "Died (STA=1)",
                  "Fitted logit line", "logit = 0 (π = 0.50)"),
       col    = c("steelblue", "tomato", "black", "gray50"),
       pch    = c(19, 19, NA, NA),
       lty    = c(NA, NA, 1, 2),
       lwd    = c(NA, NA, 2, 1.5),
       bty    = "n")

```

## Task 4h: Predicted Probabilities
```{r}
# Compute probabilities from logits
icu$prob_AGE <- exp(icu$logit_AGE) / (1 + exp(icu$logit_AGE))

# Verify probabilities saved correctly
kable(head(data.frame(
  ID       = icu$ID,
  AGE      = icu$AGE,
  STA      = icu$STA,
  Logit    = round(icu$logit_AGE, 3),
  Prob_Die = round(icu$prob_AGE,  3),
  Prob_Sur = round(1 - icu$prob_AGE, 3)), 10),
  col.names = c("ID", "AGE", "STA", "Logit",
                "P(Death)", "P(Survival)"),
  caption   = "Table 4h.1: Sample of Predicted Logits and Probabilities")

# Summary of predicted probabilities
kable(data.frame(
  Statistic = c("Min", "Q1", "Median", "Mean", "Q3", "Max"),
  Value     = round(c(min(icu$prob_AGE),
                      quantile(icu$prob_AGE, 0.25),
                      median(icu$prob_AGE),
                      mean(icu$prob_AGE),
                      quantile(icu$prob_AGE, 0.75),
                      max(icu$prob_AGE)), 3)),
  caption = "Table 4h.2: Summary of Predicted Probabilities of Death")

```

```{r}
# Sort by AGE for clean S-curve line
icu_sorted <- icu[order(icu$AGE), ]

# 1. S-curve with raw data overlaid
plot(icu$AGE, jitter(icu$STA, amount = 0.02),
     main   = "Figure 4h.1: Predicted Probability of Death by Age",
     xlab   = "Age (years)",
     ylab   = "Probability",
     pch    = 19,
     col    = ifelse(icu$STA == 1, adjustcolor("tomato", alpha.f = 0.4),
                                   adjustcolor("steelblue", alpha.f = 0.4)),
     ylim   = c(-0.1, 1.1),
     yaxt   = "n")
axis(2, at = c(0, 0.25, 0.50, 0.75, 1.0))
lines(icu_sorted$AGE, icu_sorted$prob_AGE, col = "black", lwd = 2.5)
abline(h = 0.5, lty = 2, col = "gray50", lwd = 1.5)
abline(v = age_crossover, lty = 3, col = "gray50", lwd = 1.5)
text(x = age_crossover + 2, y = 0.08,
     labels = paste0("Age = ", age_crossover, "\n(π = 0.50)"),
     cex = 0.75, col = "gray30", adj = 0)
legend("bottom",
       legend = c("Lived (STA=0)", "Died (STA=1)", "Predicted P(Death)", "π = 0.50"),
       col = c("steelblue", "tomato", "black", "gray50"),
       pch = c(19, 19, NA, NA),
       lty = c(NA, NA, 1, 2),
       lwd = c(NA, NA, 2.5, 1.5),
       bty = "n",
       horiz = TRUE,
       inset = c(0, -0.2),
       xpd = TRUE,
       cex = 0.7)


# 2. Predicted probabilities only — clean S-curve
plot(icu_sorted$AGE, icu_sorted$prob_AGE,
     type   = "l",
     lwd    = 2.5,
     col    = "black",
     main   = "Figure 4h.2: S-Shaped Logistic Curve — P(Death) by Age",
     xlab   = "Age (years)",
     ylab   = "Predicted Probability of Death",
     ylim   = c(0, 1))

# Shade the uncertainty band using binomial SE
se_prob <- sqrt(icu_sorted$prob_AGE * (1 - icu_sorted$prob_AGE) / nrow(icu))
polygon(c(icu_sorted$AGE, rev(icu_sorted$AGE)),
        c(icu_sorted$prob_AGE + 1.96 * se_prob,
          rev(icu_sorted$prob_AGE - 1.96 * se_prob)),
        col    = adjustcolor("gray70", alpha.f = 0.4),
        border = NA)
lines(icu_sorted$AGE, icu_sorted$prob_AGE, col = "black", lwd = 2.5)

abline(h   = 0.5, lty = 2, col = "gray50", lwd = 1.5)
abline(v   = age_crossover, lty = 3, col = "gray50", lwd = 1.5)

legend("top",
       legend = c("Predicted P(Death)", "95% band", "π = 0.50"),
       col    = c("black", "gray70", "gray50"),
       lty    = c(1, NA, 2),
       lwd    = c(2.5, NA, 1.5),
       pch    = c(NA, 15, NA),
       pt.cex = c(NA, 2, NA),
       bty    = "n",
       horiz = TRUE,
       inset = c(0, -0.15), xpd = TRUE)

# 3. Predicted probability of SURVIVAL (not death) — S-curve inverted
plot(icu_sorted$AGE, 1 - icu_sorted$prob_AGE,
     type   = "l",
     lwd    = 2.5,
     col    = "steelblue",
     main   = "Figure 4h.3: Predicted Probability of Survival by Age",
     xlab   = "Age (years)",
     ylab   = "Predicted Probability of Survival",
     ylim   = c(0, 1))

# Overlay raw STA (inverted: STA=0 means survived)
points(icu$AGE, jitter(1 - icu$STA, amount = 0.02),
       pch = 19,
       col = ifelse(icu$STA == 0, adjustcolor("steelblue", alpha.f = 0.3),
                                  adjustcolor("tomato",    alpha.f = 0.3)))

abline(h = 0.5, lty = 2, col = "gray50", lwd = 1.5)

legend("bottom",
       legend = c("Survived (STA=0)", "Died (STA=1)", "Predicted P(Survival)"),
       col    = c("steelblue", "tomato", "steelblue"),
       pch = c(19, 19, NA, NA),
       lty = c(NA, NA, 1, 2),
       lwd = c(NA, NA, 2.5, 1.5),
       bty = "n",
       horiz = TRUE,
       inset = c(0, -0.2),
       xpd = TRUE,
       cex = 0.7)

```

## Task 4i: Predict Probability of Survival for a Specific Age
```{r}
# Set age and predict
my_age <- 30

new_data   <- data.frame(AGE = my_age)
pred_logit <- predict(model_age, newdata = new_data, type = "link")
pred_prob_death    <- predict(model_age, newdata = new_data, type = "response")
pred_prob_survival <- 1 - pred_prob_death

cat("=== Prediction for Age", my_age, "===\n\n")
cat(sprintf("Predicted logit:               %.3f\n", pred_logit))
cat(sprintf("Predicted P(Death):            %.3f\n", pred_prob_death))
cat(sprintf("Predicted P(Survival):         %.3f\n", pred_prob_survival))

```

```{r}
# Find the AGE_CAT this age belongs to
age_cat_lookup <- cut(my_age,
                      breaks = c(15, 25, 35, 45, 55, 65, 75, 85, 95, Inf),
                      labels = c("[15,24]", "[25,34]", "[35,44]",
                                 "[45,54]", "[55,64]", "[65,74]",
                                 "[75,84]", "[85,94]", "[95+]"),
                      right  = FALSE)

# Observed proportion in that age category
obs_prop <- age_cat_summary$Proportion_Died[
              age_cat_summary$Interval == as.character(age_cat_lookup)]
obs_n    <- age_cat_summary$N[
              age_cat_summary$Interval == as.character(age_cat_lookup)]

cat(sprintf("\nAge category:                  %s\n", age_cat_lookup))
cat(sprintf("Observed N in category:        %d\n",   obs_n))
cat(sprintf("Observed P(Death) in category: %.3f\n", obs_prop))
cat(sprintf("Observed P(Survival):          %.3f\n", 1 - obs_prop))
cat(sprintf("\nModel predicted P(Survival):   %.3f\n", pred_prob_survival))
cat(sprintf("Observed P(Survival):          %.3f\n", 1 - obs_prop))
cat(sprintf("Difference:                    %.3f\n",
            pred_prob_survival - (1 - obs_prop)))
```

```{r}
# Summary comparison table
comparison <- data.frame(
  Source    = c("Logistic Model (AGE only)",
                paste0("Observed — Age Category ", age_cat_lookup),
                "Overall Sample (Task 1)"),
  P_Death   = round(c(pred_prob_death, obs_prop, mean(icu$STA)), 3),
  P_Survival = round(c(pred_prob_survival, 1 - obs_prop, 1 - mean(icu$STA)), 3)
)

kable(comparison,
      col.names = c("Source", "P(Death)", "P(Survival)"),
      caption   = paste("Table 4i.1: Predicted vs Observed Survival —",
                        "Age", my_age))
```

```{r fig.height=7, fig.width=8}
# Visualizations
# 1. S-curve with personal prediction highlighted

plot(icu_sorted$AGE, 1 - icu_sorted$prob_AGE,
     type = "l",
     lwd  = 2.5,
     col  = "steelblue",
     main = paste("Figure 4i.1: Predicted P(Survival) by Age (Age", my_age, "Highlighted)"),
     xlab = "Age (years)",
     ylab = "Predicted Probability of Survival",
     ylim = c(0, 1))
points(icu$AGE, jitter(1 - icu$STA, amount = 0.02),
       pch = 19,
       col = ifelse(icu$STA == 0, adjustcolor("steelblue", alpha.f = 0.2),
                                  adjustcolor("tomato", alpha.f = 0.2)))
abline(v = my_age, lty = 2, col = "gray40", lwd = 1.5)
abline(h = pred_prob_survival, lty = 2, col = "gray40", lwd = 1.5)
abline(h = 0.5, lty = 3, col = "gray60")
points(my_age, pred_prob_survival, pch = 18, col = "gold3", cex = 2.5)
text(x = my_age + 3, y = pred_prob_survival - 0.05,
     labels = sprintf("Age %d\nP(Surv)=%.3f", my_age, pred_prob_survival),
     cex = 0.75, col = "gray20", adj = 0)
legend("bottom",
       legend = c("Survived", "Died", "Predicted", "Your age"),
       col = c("steelblue", "tomato", "steelblue", "gold3"),
       pch = c(19, 19, NA, 18),
       lty = c(NA, NA, 1, NA),
       lwd = c(NA, NA, 2.5, NA),
       bty = "n",
       horiz = TRUE,
       inset = c(0, -0.2),
       xpd = TRUE,
       cex = 0.7)


# 2. Comparison barplot — model vs observed vs overall
barplot(t(as.matrix(comparison[, c("P_Death", "P_Survival")])),
        beside    = TRUE,
        col       = c("tomato", "steelblue"),
        border    = "white",
        main      = paste("Figure 4i.2: Survival Probability Comparison (Age", my_age, ")"),
        names.arg = c("Model", "Observed", "Overall"),
        ylab      = "Probability",
        ylim      = c(0, 1.05),
        cex.names = 0.85)
abline(h = 1 - mean(icu$STA), lty = 2, col = "gray40", lwd = 1.5)
legend("bottom",
       legend = c("P(Death)", "P(Survival)", "Overall P(Survival)"),
       col = c("tomato", "steelblue", "gray40"),
       pch = c(15, 15, NA),
       lty = c(NA, NA, 2),
       lwd = c(NA, NA, 1.5),
       bty = "n",
       horiz = TRUE,
       inset = c(0, -0.22),
       xpd = TRUE,
       cex = 0.75)


```

# Task 5: Recommendations for Next Phase of Modeling
## 5a. Summary of Key Findings So Far
```{r}
# Compile evidence gathered across Tasks 1-4 into a summary table
findings <- data.frame(
  Task     = c("Task 1", "Task 2", "Task 2", "Task 3", "Task 3", "Task 4"),
  Variable = c("Overall Sample", "SEX", "SEX", "TYP", "TYP", "AGE"),
  Finding  = c(
    paste0("20% mortality rate (", sum(icu$STA == 1),
           " deaths / 200 patients). Class imbalance noted."),
    paste0("P(Survival | Male) = ",   round(prob_male_survive,   3),
           ", P(Survival | Female) = ", round(prob_female_survive, 3)),
    paste0("OR (Male vs Female) = ",  round(odds_male / odds_female, 3),
           " — minimal difference between sexes"),
    paste0("P(Survival | Elective) = ", round(prob_elective_survive,  3),
           ", P(Survival | Emergency) = ", round(prob_emergency_survive, 3)),
    paste0("OR (Elective vs Emergency) = ", round(odds_elective / odds_emergency, 3),
           " — strong predictor of survival"),
    paste0("β₁ = ", round(coef(model_age)[2], 3),
           ", OR per year = ", round(exp(coef(model_age)[2]), 3),
           ", p = ", round(summary(model_age)$coefficients[2, 4], 3),
           " — significant but modest effect")
  )
)

kable(findings,
      col.names = c("Task", "Variable", "Key Finding"),
      caption   = "Table 5.1: Summary of Key Findings Across Tasks 1-4")
```

## 5b. Univariate Screening — all candidate predictors
```{r}
# Screen all binary/continuous predictors individually to identify which warrant inclusion in a multivariable model

predictors <- c("AGE", "SEX", "RACE", "SER", "CAN", "CRN", "INF",
                "CPR", "SYS", "HRA", "PRE", "TYP", "FRA",
                "PO2", "PH", "PCO", "BIC", "CRE", "LOC")

screening_results <- do.call(rbind, lapply(predictors, function(var) {
  formula  <- as.formula(paste("STA ~", var))
  fit      <- glm(formula, data = icu, family = binomial)
  coef_row <- summary(fit)$coefficients

  # Handle multi-level variables (RACE, LOC) — report omnibus deviance test
  null_dev <- fit$null.deviance
  res_dev  <- fit$deviance
  g_stat   <- round(null_dev - res_dev, 3)
  df_diff  <- fit$df.null - fit$df.residual
  p_lrt    <- round(1 - pchisq(g_stat, df_diff), 3)

  # For single-predictor variables grab the slope row
  if (nrow(coef_row) == 2) {
    data.frame(
      Predictor = var,
      Estimate  = round(coef_row[2, 1], 3),
      OR        = round(exp(coef_row[2, 1]), 3),
      z         = round(coef_row[2, 3], 3),
      p_Wald    = round(coef_row[2, 4], 3),
      G_stat    = g_stat,
      p_LRT     = p_lrt,
      AIC       = round(AIC(fit), 3)
    )
  } else {
    # Multi-level: report LRT only
    data.frame(
      Predictor = var,
      Estimate  = NA,
      OR        = NA,
      z         = NA,
      p_Wald    = NA,
      G_stat    = g_stat,
      p_LRT     = p_lrt,
      AIC       = round(AIC(fit), 3)
    )
  }
}))

kable(screening_results,
      col.names = c("Predictor", "Estimate", "OR", "z",
                    "p (Wald)", "G Statistic", "p (LRT)", "AIC"),
      caption   = "Table 5.2: Univariate Screening — All Candidate Predictors",
      na        = "—")
```

## 5c. Visualizations
```{r fig.height=7, fig.width=8}
# 1. Dot plot of LRT p-values — which predictors are significant?
screening_results$Predictor <- factor(screening_results$Predictor,
                                       levels = screening_results$Predictor[
                                         order(screening_results$p_LRT)])

dotchart(screening_results$p_LRT,
         labels = screening_results$Predictor,
         main   = "Figure 5.1: Univariate LRT p-values by Predictor\n(Sorted Ascending)",
         xlab   = "LRT p-value",
         pch    = 19,
         col    = ifelse(screening_results$p_LRT <= 0.05, "tomato", "steelblue"),
         lcolor = "gray90")
abline(v   = 0.05, lty = 2, col = "gray40", lwd = 1.5)
text(x      = 0.05,
     y      = 1,
     labels = "α = 0.05",
     cex    = 0.8,
     col    = "gray30",
     adj    = -0.1)

# 2. AIC comparison across univariate models
screening_sorted_aic <- screening_results[order(screening_results$AIC), ]

barplot(screening_sorted_aic$AIC,
        names.arg = screening_sorted_aic$Predictor,
        col       = ifelse(screening_sorted_aic$p_LRT <= 0.05, "tomato", "steelblue"),
        border    = "white",
        main      = "Figure 5.2: AIC by Univariate Model",
        xlab      = "",
        ylab      = "AIC",
        las       = 2,
        cex.names = 0.75,
        ylim      = c(0, max(screening_sorted_aic$AIC) * 1.1))
title(xlab = "Predictor", line = 5.5)
abline(h = AIC(model_age), lty = 2, col = "gray40", lwd = 1.5)
text(x = 2, y = AIC(model_age) + 3,
     labels = paste0("AGE AIC = ", round(AIC(model_age), 3)),
     cex = 0.75, col = "gray30", adj = 0)
legend("bottom",
       legend = c("Significant (p≤0.05)", "Not significant"),
       fill = c("tomato", "steelblue"),
       border = "white",
       bty = "n",
       horiz = TRUE,
       inset = c(0, -0.25),
       xpd = TRUE,
       cex = 0.75)

# 3. OR forest plot for significant predictors
sig_predictors <- screening_results[
  !is.na(screening_results$p_Wald) & screening_results$p_LRT <= 0.05, ]

if (nrow(sig_predictors) > 0) {
  par(mar = c(5, 6, 4, 2))
  plot(sig_predictors$OR,
       1:nrow(sig_predictors),
       pch   = 18,
       cex   = 1.8,
       col   = "tomato",
       xlim  = c(0, max(sig_predictors$OR, na.rm = TRUE) * 1.2),
       xlab  = "Odds Ratio",
       ylab  = "",
       yaxt  = "n",
       main  = "Figure 5.3: Odds Ratios — Significant Univariate Predictors")
  axis(2,
       at     = 1:nrow(sig_predictors),
       labels = sig_predictors$Predictor,
       las    = 1)
  abline(v   = 1,   lty = 2, col = "gray40", lwd = 1.5)
  abline(h   = 1:nrow(sig_predictors), lty = 3, col = "gray85")
  par(mar = c(5, 4, 4, 2))
}
```

```{r}

```