---
title: "Assignment 6 - Finalizing the Ames Housing Price Prediction Model"
author: "Shishir Deshpande"
date: "February 15, 2026"
output:
  html_document: default
  word_document: default
  pdf_document: default
---


# Importing required libraries
```{r}
library(tidyr)
library(readr)
library(knitr)
library(corrplot)
library(dplyr)
library(MASS)
library(car)

# Prevent namespace conflicts
select <- dplyr::select
filter <- dplyr::filter
```

# Importing the dataset
```{r}
data <- read.csv("ames_housing_data.csv")

cat("The dataset has ", nrow(data), "rows and ", ncol(data), "columns")

summary_matrix <- summary(data %>% select(where(is.numeric)))
summary_df <- as.data.frame(summary_matrix)
summary_df <- summary_df %>%
  separate(Freq, into = c("Stat", "Value"), sep = ":", extra = "merge") %>%
  mutate(Value = trimws(Value)) %>%
  pivot_wider(names_from = Stat, values_from = Value)

kable(summary_df)

```

# Preparatory work
## Sampling
```{r}
### We will define the sample population similar to Modeling Assignment 1. 
## Remove non normal sale conditions
## Filter to single family homes
ames_data_clean <- data %>%
  filter(
    BldgType == "1Fam",
    SaleCondition == "Normal", 
    GrLivArea <= 4000)

cat("The cleaned dataset (aka Sample Population) has ", nrow(ames_data_clean), "rows and ", ncol(ames_data_clean), "columns")
```

## Handling missing values
```{r}

## Drop variables with >50% missing values and the indicator variables (not useful for modeling); impute the median for LotFrontage since it has 20%~ missing values and is quantitative/continuous; the rest will be filled with 0/None as applicable. Having a zero/None is also valuable as the lack of something can also change sale prices.
ames_data_clean <- ames_data_clean %>%
  select(-PoolQC, -MiscFeature, -Alley, -Fence, -FireplaceQu, -SID, -PID) %>%
  mutate(
    LotFrontage = replace_na(LotFrontage, median(LotFrontage, na.rm = TRUE)),
    MasVnrArea = replace_na(MasVnrArea, median(MasVnrArea, na.rm = TRUE)),
    across(c(GarageFinish, GarageType, GarageQual, GarageCond, BsmtExposure, BsmtFinType2, 
             BsmtQual, BsmtCond, BsmtFinType1, MasVnrType, Electrical), ~replace_na(., "None")),
    across(c(GarageYrBlt, BsmtFullBath, BsmtHalfBath), ~replace_na(., 0))
  )

cat("The cleaned dataset (aka Sample Population) has ", nrow(ames_data_clean), "rows and ", ncol(ames_data_clean), "columns")
```

## Selecting categorical variables based purely on mean difference between levels (the greater the difference, the more predictive power).
```{r}
# Identify all categorical variables
cat_vars <- names(ames_data_clean)[sapply(ames_data_clean, function(x) is.character(x) | is.factor(x))]

cat(paste(cat_vars, collapse = ", "))

```

**High Priority (Strong Predictors):**
- ExterQual (Exterior Quality) - Overall material and finish quality directly impacts value
- KitchenQual - Kitchen quality is a major selling point and renovation cost driver
- BsmtQual (Basement Quality) - Finished basement quality adds significant living space value
- HeatingQC (Heating Quality) - Essential comfort feature affecting desirability

**Medium Priority (Moderate Predictors):**
- Foundation - Type of foundation affects durability and maintenance costs
- GarageType - Attached vs detached vs none significantly impacts convenience and value
- GarageFinish - Finished garages add functional space
- BsmtExposure - Walkout basements are more valuable than enclosed ones
- MasVnrType - Brick/stone veneer adds curb appeal and perceived quality

**Lower Priority (Weaker Predictors):**
- RoofStyle - Aesthetic preference with some functional differences
- Exterior1/Exterior2 - Siding material affects maintenance and appearance
- SaleType - How property was sold may correlate with price
- LotShape - Regular lots typically more desirable than irregular
- Zoning - Residential classifications affect permitted uses
- SaleCondition - Not relevant as we have only type post our sampling filter

*I will exclude 'Neighborhood' because the assignment's drop list removes it before variable selection, likely due to its 25+ levels creating too many dummy variables and potential overfitting concerns.*


# Based on this, I will analyze the high priority categorical variables (ExterQual, KitchenQual, BsmtQual, HeatingQC)
```{r}
# High priority categorical variables
priority_cats <- c("ExterQual", "KitchenQual", "BsmtQual", "HeatingQC")

# Summary statistics by categorical variable levels
cat_summary <- data.frame()
for(var in priority_cats) {
  stats <- ames_data_clean %>%
    group_by(!!sym(var)) %>%
    summarise(N = n(), Mean = mean(SalePrice), Median = median(SalePrice), 
              SD = sd(SalePrice)) %>%
    arrange(desc(Mean))
  
  mean_diff <- max(stats$Mean) - min(stats$Mean)
  cat_summary <- rbind(cat_summary, data.frame(Variable = var, MeanDiff = mean_diff))
  
  cat("\n", var, ":\n")
  print((kable(stats)))
}

# Rank by mean difference
cat_summary <- cat_summary %>% arrange(desc(MeanDiff))
cat("\n\nVariables Ranked by Mean Difference:\n")
kable(cat_summary)

cat("\n\nWhy mean difference matters: Large mean differences indicate strong predictive power - the variable clearly separates high-priced from low-priced homes, making it valuable for the model.\n")

```

## Deciding which categorical variables to convert to dummy and retain for modeling
```{r}
# Fit regression models to calculate R-squared for each categorical variable
rsq_results <- data.frame()
for(var in priority_cats) {
  model <- lm(SalePrice ~ get(var), data = ames_data_clean)
  rsq <- summary(model)$r.squared
  num_levels <- length(unique(ames_data_clean[[var]]))
  rsq_results <- rbind(rsq_results, data.frame(
    Variable = var, 
    R_squared = rsq, 
    NumLevels = num_levels,
    DummiesNeeded = num_levels - 1
  ))
}

rsq_results <- rsq_results %>% arrange(desc(R_squared))
kable(rsq_results)

# Select variables with R² > 0.25 (explain more than 25% of price variation)
selected_cats <- rsq_results$Variable[rsq_results$R_squared > 0.25]
cat("\nSelected variables (R² > 0.25):", paste(selected_cats, collapse = ", "), "\n")

# Create dummy variables
total_dummies <- 0
for(var in selected_cats) {
  # Use model.matrix with intercept to get all-but-one dummies automatically
  # Reference level (first alphabetically) is excluded - this is the controlled choice
  dummies <- model.matrix(~ factor(ames_data_clean[[var]]))[, -1]
  # Name columns: strip the "factor(ames_data_clean[[var]])" prefix
  colnames(dummies) <- paste0(var, "_", levels(factor(ames_data_clean[[var]]))[-1])
  ames_data_clean <- cbind(ames_data_clean, dummies)
  total_dummies <- total_dummies + ncol(dummies)
}

cat("Total dummy variables created:", total_dummies, "\n")

```

# Predictive Modeling
```{r}
# Set the seed on the random number generator so you get the same split every time that you run the code.
set.seed(123)
ames_data_clean$u <- runif(n = dim(ames_data_clean)[1], min=0, max=1);

# Define these two variables for later use;
ames_data_clean$QualityIndex <- ames_data_clean$OverallQual * ames_data_clean$OverallCond;
ames_data_clean$TotalSqftCalc <- ames_data_clean$BsmtFinSF1 + ames_data_clean$BsmtFinSF2 + ames_data_clean$GrLivArea;

# Create train/test split;
train.df <- subset(ames_data_clean, u < 0.70);
test.df  <- subset(ames_data_clean, u >= 0.70);

# Check your data split. The sum of the parts should equal the whole.
# Do your totals add up?
cat("The original dataset has", dim(ames_data_clean)[1], " records.")
dim(train.df)[1]
dim(test.df)[1]
cat("The data has ", (dim(train.df)[1]), "training records and ", (dim(test.df)[1]), "test records")

```

```{r}
# Train/Test partition table
kable(data.frame(
  Dataset = c("Training", "Test", "Total"),
  Observations = c(nrow(train.df), nrow(test.df), nrow(ames_data_clean)),
  Percentage = paste0(round(c(nrow(train.df), nrow(test.df), nrow(ames_data_clean)) / nrow(ames_data_clean) * 100, 1), "%")
))

```

```{r}
# Drop unnecessary columns
drop.list <- c('SID','PID','LotConfig','Neighborhood','HouseStyle','YearBuilt','YearRemodel',
'Exterior1','BsmtFinSF1','BsmtFinSF2','CentralAir','YrSold','MoSold','SaleCondition','u','train','I2010','BsmtFullBath','BsmtHalfBath','FullBath','HalfBath',
	'FireplaceInd1','FireplaceInd2','OverallQual','OverallCond','PoolArea','GrLivArea');

train.clean <- train.df[,!(names(train.df) %in% c(drop.list, priority_cats))]
test.clean <- test.df[,!(names(test.df) %in% c(drop.list, priority_cats))]

```

```{r}
# Keep only numeric columns
train.clean <- train.clean %>% select(where(is.numeric))
test.clean <- test.clean %>% select(where(is.numeric))

head(train.clean)
cat("Number of predictor variables in pool:", ncol(train.clean) - 1, "\n")
```

```{r}
library(kableExtra)

# Define the upper model as the FULL model
upper.lm <- lm(SalePrice ~ ., data = train.clean);
summary(upper.lm)

# Define the lower model as the Intercept model
lower.lm <- lm(SalePrice ~ 1, data=train.clean);

# Need a SLR to initialize stepwise selection
sqft.lm <- lm(SalePrice ~ TotalSqftCalc, data=train.clean);
summary(sqft.lm)

```

```{r}
# Note: There is only one function for classical model selection in R - stepAIC();
# stepAIC() is part of the MASS library.
# The MASS library comes with the BASE R distribution, but you still need to load it;

# Call stepAIC() for variable selection
forward.lm <- stepAIC(object=lower.lm,scope=list(upper=formula(upper.lm),lower=~1),
direction=c('forward'));
summary(forward.lm)

backward.lm <- stepAIC(object=upper.lm,direction=c('backward'));
summary(backward.lm)

stepwise.lm <- stepAIC(object=sqft.lm,scope=list(upper=formula(upper.lm),lower=~1),
direction=c('both'));
summary(stepwise.lm)

```

```{r}
junk.lm <- lm(SalePrice ~ OverallQual + OverallCond + QualityIndex + GrLivArea + TotalSqftCalc, data=train.df)
summary(junk.lm)

```

```{r}
# Load required libraries
# ========================================
# UPPER MODEL (Full Model)
# ========================================
cat("

## UPPER MODEL (Full Model with All 36 Predictors)

")

# Coefficient table
cat("
### Coefficient Table:

")
kable(summary(upper.lm)$coefficients,
      digits = 4,
      caption = "Coefficients for Upper Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ANOVA table
cat("
### ANOVA Table:

")
kable(anova(upper.lm),
      digits = 4,
      caption = "ANOVA for Upper Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Model summary
cat("
### Model Summary:
")
cat("Number of Predictors:", length(coef(upper.lm)) - 1, "
")
cat("Adjusted R-squared:", round(summary(upper.lm)$adj.r.squared, 4), "
")
cat("AIC:", round(AIC(upper.lm), 2), "
")
cat("BIC:", round(BIC(upper.lm), 2), "

")


# ========================================
# SLR MODEL (Simple Linear Regression)
# ========================================
cat("

## SLR MODEL (SalePrice ~ TotalSqftCalc)

")

# Coefficient table
cat("
### Coefficient Table:

")
kable(summary(sqft.lm)$coefficients,
      digits = 4,
      caption = "Coefficients for SLR Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ANOVA table
cat("
### ANOVA Table:

")
kable(anova(sqft.lm),
      digits = 4,
      caption = "ANOVA for SLR Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Model summary
cat("
### Model Summary:
")
cat("Number of Predictors:", length(coef(sqft.lm)) - 1, "
")
cat("Adjusted R-squared:", round(summary(sqft.lm)$adj.r.squared, 4), "
")
cat("AIC:", round(AIC(sqft.lm), 2), "
")
cat("BIC:", round(BIC(sqft.lm), 2), "

")


# ========================================
# FORWARD SELECTION MODEL
# ========================================
cat("

## FORWARD SELECTION MODEL

")

# Coefficient table
cat("
### Coefficient Table:

")
kable(summary(forward.lm)$coefficients,
      digits = 4,
      caption = "Coefficients for Forward Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ANOVA table
cat("
### ANOVA Table:

")
kable(anova(forward.lm),
      digits = 4,
      caption = "ANOVA for Forward Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Model summary
cat("
### Model Summary:
")
cat("Number of Predictors:", length(coef(forward.lm)) - 1, "
")
cat("Adjusted R-squared:", round(summary(forward.lm)$adj.r.squared, 4), "
")
cat("AIC:", round(AIC(forward.lm), 2), "
")
cat("BIC:", round(BIC(forward.lm), 2), "

")


# ========================================
# BACKWARD SELECTION MODEL
# ========================================
cat("

## BACKWARD SELECTION MODEL

")

# Coefficient table
cat("
### Coefficient Table:

")
kable(summary(backward.lm)$coefficients,
      digits = 4,
      caption = "Coefficients for Backward Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ANOVA table
cat("
### ANOVA Table:

")
kable(anova(backward.lm),
      digits = 4,
      caption = "ANOVA for Backward Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Model summary
cat("
### Model Summary:
")
cat("Number of Predictors:", length(coef(backward.lm)) - 1, "
")
cat("Adjusted R-squared:", round(summary(backward.lm)$adj.r.squared, 4), "
")
cat("AIC:", round(AIC(backward.lm), 2), "
")
cat("BIC:", round(BIC(backward.lm), 2), "

")


# ========================================
# STEPWISE SELECTION MODEL
# ========================================
cat("

## STEPWISE SELECTION MODEL

")

# Coefficient table
cat("
### Coefficient Table:

")
kable(summary(stepwise.lm)$coefficients,
      digits = 4,
      caption = "Coefficients for Stepwise Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ANOVA table
cat("
### ANOVA Table:

")
kable(anova(stepwise.lm),
      digits = 4,
      caption = "ANOVA for Stepwise Selection Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Model summary
cat("
### Model Summary:
")
cat("Number of Predictors:", length(coef(stepwise.lm)) - 1, "
")
cat("Adjusted R-squared:", round(summary(stepwise.lm)$adj.r.squared, 4), "
")
cat("AIC:", round(AIC(stepwise.lm), 2), "
")
cat("BIC:", round(BIC(stepwise.lm), 2), "

")


```

```{r}
# Compute the VIF values
kable(sort(vif(forward.lm),decreasing=TRUE))
kable(sort(vif(backward.lm),decreasing=TRUE))
kable(sort(vif(stepwise.lm),decreasing=TRUE))
kable(sort(vif(junk.lm),decreasing=TRUE))

```

```{r}
# Compare the three variable selection models
cat("\n=== MODEL COMPARISON ===\n")
cat("Forward Selection - Number of variables:", length(coef(forward.lm)) - 1, "\n")
cat("Backward Selection - Number of variables:", length(coef(backward.lm)) - 1, "\n")
cat("Stepwise Selection - Number of variables:", length(coef(stepwise.lm)) - 1, "\n")
cat("Junk Model - Number of variables:", length(coef(junk.lm)) - 1, "\n\n")

# Check if models are identical
same_forward_backward <- setequal(names(coef(forward.lm)), names(coef(backward.lm)))
same_forward_stepwise <- setequal(names(coef(forward.lm)), names(coef(stepwise.lm)))
same_backward_stepwise <- setequal(names(coef(backward.lm)), names(coef(stepwise.lm)))

cat("Are Forward and Backward the same?", same_forward_backward, "\n")
cat("Are Forward and Stepwise the same?", same_forward_stepwise, "\n")
cat("Are Backward and Stepwise the same?", same_backward_stepwise, "\n")

```

```{r}
# Calculate in-sample fit metrics for all models
calc_metrics <- function(model, data) {
  n <- nrow(data)
  k <- length(coef(model)) - 1
  mse <- mean(model$residuals^2)
  mae <- mean(abs(model$residuals))
  mape <- mean(abs(model$residuals / data$SalePrice)) * 100
  
  return(c(
    Adj_R2 = summary(model)$adj.r.squared,
    AIC = AIC(model),
    BIC = BIC(model),
    MSE = mse,
    MAE = mae,
    MAPE = mape
  ))
}

# Calculate for all models
forward_metrics <- calc_metrics(forward.lm, train.clean)
backward_metrics <- calc_metrics(backward.lm, train.clean)
stepwise_metrics <- calc_metrics(stepwise.lm, train.clean)
junk_metrics <- calc_metrics(junk.lm, train.df)

# Combine into table
metrics_table <- data.frame(
  Forward = forward_metrics,
  Backward = backward_metrics,
  Stepwise = stepwise_metrics,
  Junk = junk_metrics
)

# Calculate rankings (1 = best)
rank_table <- metrics_table
rank_table[1,] <- rank(-metrics_table[1,], ties.method = "first")  # Higher Adj R2 is better
rank_table[2:6,] <- apply(metrics_table[2:6,], 1, rank, ties.method = "first")  # Lower is better

cat("\n=== IN-SAMPLE MODEL COMPARISON ===\n")
kable(metrics_table, digits = 2, caption = "Training Data Fit Metrics")
cat("\n")
kable(rank_table, caption = "Model Rankings (1 = Best)")

```

```{r}
# Generate predictions on test data
forward.test <- predict(forward.lm, newdata = test.clean)
backward.test <- predict(backward.lm, newdata = test.clean)
stepwise.test <- predict(stepwise.lm, newdata = test.clean)
junk.test <- predict(junk.lm, newdata = test.df)

# Calculate test MSE, MAE, and MAPE
test_metrics <- data.frame(
  Forward = c(
    MSE = mean((test.clean$SalePrice - forward.test)^2),
    MAE = mean(abs(test.clean$SalePrice - forward.test)),
    MAPE = mean(abs((test.clean$SalePrice - forward.test) / test.clean$SalePrice)) * 100
  ),
  Backward = c(
    MSE = mean((test.clean$SalePrice - backward.test)^2),
    MAE = mean(abs(test.clean$SalePrice - backward.test)),
    MAPE = mean(abs((test.clean$SalePrice - backward.test) / test.clean$SalePrice)) * 100
  ),
  Stepwise = c(
    MSE = mean((test.clean$SalePrice - stepwise.test)^2),
    MAE = mean(abs(test.clean$SalePrice - stepwise.test)),
    MAPE = mean(abs((test.clean$SalePrice - stepwise.test) / test.clean$SalePrice)) * 100
  ),
  Junk = c(
    MSE = mean((test.df$SalePrice - junk.test)^2),
    MAE = mean(abs(test.df$SalePrice - junk.test)),
    MAPE = mean(abs((test.df$SalePrice - junk.test) / test.df$SalePrice)) * 100
  )
)

# Add rankings
test_ranks <- apply(test_metrics, 1, rank)

cat("\n=== OUT-OF-SAMPLE PREDICTIVE ACCURACY ===\n")
kable(test_metrics, digits = 2, caption = "Test Data Performance")
cat("\n")
kable(t(test_ranks), caption = "Test Rankings (1 = Best)")

# Compare train vs test MSE
cat("\n\nTrain vs Test MSE:\n")
comparison <- rbind(
  Train = metrics_table["MAPE",],
  Test = test_metrics["MAPE",],
  Difference = test_metrics["MAPE",] - metrics_table["MAPE",]
)
kable(comparison, digits = 2)

```

```{r}
# Function to assign prediction grades
assign_grade <- function(pct_error) {
  ifelse(pct_error <= 0.10, 'Grade 1: [0,0.10]',
         ifelse(pct_error <= 0.15, 'Grade 2: (0.10,0.15]',
                ifelse(pct_error <= 0.25, 'Grade 3: (0.15,0.25]',
                       'Grade 4: (0.25+]')))
}

# Training data prediction grades
forward.trainPCT <- abs(forward.lm$residuals) / train.clean$SalePrice
backward.trainPCT <- abs(backward.lm$residuals) / train.clean$SalePrice
stepwise.trainPCT <- abs(stepwise.lm$residuals) / train.clean$SalePrice
junk.trainPCT <- abs(junk.lm$residuals) / train.df$SalePrice

# Test data prediction grades
forward.testPCT <- abs(test.clean$SalePrice - forward.test) / test.clean$SalePrice
backward.testPCT <- abs(test.clean$SalePrice - backward.test) / test.clean$SalePrice
stepwise.testPCT <- abs(test.clean$SalePrice - stepwise.test) / test.clean$SalePrice
junk.testPCT <- abs(test.df$SalePrice - junk.test) / test.df$SalePrice

# Create grade distributions (cleaner format)
train_grade_summary <- rbind(
  Forward = prop.table(table(assign_grade(forward.trainPCT))),
  Backward = prop.table(table(assign_grade(backward.trainPCT))),
  Stepwise = prop.table(table(assign_grade(stepwise.trainPCT))),
  Junk = prop.table(table(assign_grade(junk.trainPCT)))
)

test_grade_summary <- rbind(
  Forward = prop.table(table(assign_grade(forward.testPCT))),
  Backward = prop.table(table(assign_grade(backward.testPCT))),
  Stepwise = prop.table(table(assign_grade(stepwise.testPCT))),
  Junk = prop.table(table(assign_grade(junk.testPCT)))
)

cat("\n=== OPERATIONAL VALIDATION ===\n")
cat("\nTraining Data Prediction Grades:\n")
kable(train_grade_summary, digits = 3)

cat("\n\nTest Data Prediction Grades:\n")
kable(test_grade_summary, digits = 3)

# Underwriting quality check (>50% Grade 1)
cat("\n\nUnderwriting Quality (Grade 1 %):\n")
underwriting <- rbind(
  Train = train_grade_summary[, 1],
  Test = test_grade_summary[, 1]
)
kable(underwriting, digits = 3)

cat("\n\nModels meeting underwriting quality (>0.50):\n")
print(colnames(underwriting)[underwriting["Test",] > 0.50])

```

```{r}
# Select best model based on test performance
best_idx <- which.min(test_metrics["MSE",])
best_name <- names(best_idx)
best.lm <- get(paste0(tolower(best_name), ".lm"))

cat("\n=== BEST MODEL SELECTION ===\n")
cat("Best Model:", best_name, "\n")
cat("Test MSE:", round(test_metrics["MSE", best_idx], 2), "\n")
cat("Test MAPE:", round(test_metrics["MAPE", best_idx], 2), "%\n")
cat("Grade 1 (Test):", round(test_grade_summary[best_name, 1], 3), "\n\n")

# Display model summary
cat("Model Variables:\n")
print(names(coef(best.lm)))
cat("\n\nModel Summary:\n")
print(summary(best.lm))

```

```{r}
cat("\n=== COEFFICIENT SIGN VERIFICATION ===\n\n")

# Extract coefficients (excluding intercept)
coef_df <- data.frame(
  Variable = names(coef(best.lm))[-1],
  Coefficient = coef(best.lm)[-1]
) %>%
  mutate(Direction = ifelse(Coefficient > 0, "Positive", "Negative"))

# Reference category for all quality dummies is Excellent (Ex)
# So all other levels should be negative (discount vs Excellent)
expected_signs <- c(
  TotalSqftCalc  = "Positive",  # More space = higher price
  BsmtUnfSF      = "Positive",  # More unfinished basement still adds space
  QualityIndex   = "Positive",  # Higher quality/condition = higher price
  LotArea        = "Positive",  # Larger lot = higher price
  MasVnrArea     = "Positive",  # Masonry veneer adds curb appeal
  Fireplaces     = "Positive",  # Amenity adds value
  GarageCars     = "Positive",  # More garage capacity = higher price
  GarageArea     = "Positive",  # Larger garage = higher price
  SecondFlrSF    = "Positive",  # More living space = higher price
  WoodDeckSF     = "Positive",  # Outdoor amenity adds value
  ScreenPorch    = "Positive",  # Outdoor amenity adds value
  LotFrontage    = "Positive",  # More street frontage = higher price
  # Quality dummies — all negative vs Excellent reference
  BsmtQual_Gd   = "Negative",  # Good < Excellent
  BsmtQual_TA   = "Negative",  # Typical < Excellent
  BsmtQual_Fa   = "Negative",  # Fair < Excellent
  BsmtQual_Po   = "Negative",  # Poor < Excellent
  BsmtQual_None = "Negative",  # No basement < Excellent
  ExterQual_Gd  = "Negative",  # Good < Excellent
  ExterQual_TA  = "Negative",  # Typical < Excellent
  ExterQual_Fa  = "Negative",  # Fair < Excellent
  KitchenQual_Gd = "Negative", # Good < Excellent
  KitchenQual_TA = "Negative", # Typical < Excellent
  KitchenQual_Fa = "Negative", # Fair < Excellent
  KitchenQual_Po = "Negative", # Poor < Excellent
  # Ambiguous — theory doesn't give a clear single direction
  BedroomAbvGr  = "Ambiguous", # More bedrooms can reduce price per room
  KitchenAbvGr  = "Negative",  # Multiple kitchens signals conversion/unusual
  EnclosedPorch = "Negative",  # Less desirable than open porch
  GarageYrBlt   = "Ambiguous", # Newer = better but collinear with house age
  LowQualFinSF  = "Negative",  # Low quality finish drags price down
  ThreeSsnPorch = "Ambiguous", # Niche feature, market dependent
  SubClass      = "Ambiguous"  # Numeric code, direction not straightforward
)

# Join expected to actual
coef_df <- coef_df %>%
  mutate(
    Expected = expected_signs[Variable],
    Flag = case_when(
      Expected == "Ambiguous" ~ "Review",
      Direction == Expected   ~ "OK",
      TRUE                    ~ "WARNING"
    )
  )

kable(coef_df, digits = 2, row.names = FALSE,
      caption = "Coefficient Sign Verification")

# Summary
n_warn <- sum(coef_df$Flag == "WARNING")
n_ok   <- sum(coef_df$Flag == "OK")
n_rev  <- sum(coef_df$Flag == "Review")

cat("\nSummary:\n")
cat("• Signs as expected:", n_ok, "variables\n")
cat("• Flagged for review (ambiguous theory):", n_rev, "variables\n")
cat("• Unexpected signs:", n_warn, "variables\n")

if(n_warn > 0) {
  cat("\nVariables with unexpected signs:\n")
  print(coef_df %>% filter(Flag == "WARNING") %>% 
        select(Variable, Coefficient, Expected))
}
```

```{r}
# Residual diagnostics for best model
cat("\n=== RESIDUAL DIAGNOSTICS ===\n\n")

# Diagnostic plots
par(mfrow = c(2, 2))
plot(best.lm)
par(mfrow = c(1, 1))

# Shapiro-Wilk test (sample 5000 for large datasets)
if(length(best.lm$residuals) > 5000) {
  shapiro_result <- shapiro.test(sample(best.lm$residuals, 5000))
} else {
  shapiro_result <- shapiro.test(best.lm$residuals)
}

cat("\nShapiro-Wilk Normality Test:\n")
cat("W =", round(shapiro_result$statistic, 4), ", p-value =", 
    format.pval(shapiro_result$p.value, digits = 3), "\n")

# Breusch-Pagan test for heteroscedasticity
library(lmtest)
bp_result <- bptest(best.lm)
cat("\nBreusch-Pagan Heteroscedasticity Test:\n")
cat("BP =", round(bp_result$statistic, 4), ", p-value =", 
    format.pval(bp_result$p.value, digits = 3), "\n")

# Cook's Distance for influential observations
cooks_d <- cooks.distance(best.lm)
influential <- which(cooks_d > 0.5)
cat("\nInfluential Observations (Cook's D > 0.5):", 
    ifelse(length(influential) > 0, length(influential), "None"), "\n")

```

```{r}
cat("\n=== HETEROSCEDASTICITY ASSESSMENT ===\n\n")

cat("The Breusch-Pagan test confirms significant heteroscedasticity",
    "(BP =", round(bp_result$statistic, 2), 
    ", p =", format.pval(bp_result$p.value, digits = 3), ").\n\n")

cat("Implications:\n")
cat("• OLS coefficient estimates remain unbiased but are no longer efficient\n")
cat("• Standard errors are unreliable, making hypothesis tests and p-values invalid\n")
cat("• Prediction intervals will be miscalibrated (too narrow for high-value homes)\n\n")

cat("Why transformation is deferred:\n")
cat("• A log(SalePrice) model would require all predictions and metrics to be\n")
cat("  back-transformed via exp(), complicating MAPE and grade calculations\n")
cat("• The assignment's operational validation framework (MAPE, prediction grades)\n")
cat("  is defined on raw SalePrice scale — back-transformation introduces\n")
cat("  its own bias via Jensen's inequality\n")
cat("• The Residuals vs Fitted plot shows the heteroscedasticity is concentrated\n")
cat("  in the upper tail (high-value homes), which is a known feature of housing\n")
cat("  data rather than a modelling error\n")
cat("• Predictive accuracy on the test set (MAPE =", 
    round(test_metrics["MAPE", best_name], 2), "%, Grade 1 =",
    round(test_grade_summary[best_name, 1] * 100, 1), 
    "%) meets the GSE underwriting quality standard despite heteroscedasticity\n\n")

cat("Recommended next steps (out of scope for this assignment):\n")
cat("• Refit using log(SalePrice) as response and compare BP statistic\n")
cat("• Alternatively, apply weighted least squares (WLS) with weights\n")
cat("  proportional to 1/fitted values\n")
```

```{r}
# R-squared change analysis
cat("\n=== R-SQUARED CHANGE ANALYSIS ===\n")
full_r2 <- summary(best.lm)$r.squared
vars <- names(coef(best.lm))[-1]

rsq_change <- sapply(vars, function(v) {
  formula_str <- paste("SalePrice ~", paste(setdiff(vars, v), collapse = " + "))
  reduced <- lm(as.formula(formula_str), data = train.clean)
  full_r2 - summary(reduced)$r.squared
})

rsq_df <- data.frame(
  Variable = names(rsq_change),
  R2_Change = rsq_change,
  Pct_Change = (rsq_change / full_r2) * 100
) %>% arrange(R2_Change)

cat("\nVariables ranked by R² contribution:\n")
kable(rsq_df, digits = 4, row.names = FALSE)

# Identify weak contributors
weak_vars <- rsq_df$Variable[rsq_df$R2_Change < 0.001]
cat("\n\nWeak contributors (R² change < 0.001):\n")
print(weak_vars)


```

```{r}
# Get remaining variables
strong_vars <- setdiff(names(coef(best.lm))[-1], weak_vars)

# Build trimmed model formula
trimmed_formula <- as.formula(paste("SalePrice ~", paste(strong_vars, collapse = " + ")))

# Fit trimmed model
trimmed.lm <- lm(trimmed_formula, data = train.clean)

summary(trimmed.lm)
```

```{r}
# Compare full vs trimmed model
trimmed.test <- predict(trimmed.lm, newdata = test.clean)

comparison_table <- data.frame(
  Metric = c("Predictors", "Adj R²", "AIC", "BIC", "Train MAPE (%)", "Test MAPE (%)","Grade 1 Test (%)"),
  Full = c(
    length(coef(best.lm)) - 1,
    round(summary(best.lm)$adj.r.squared, 4),
    round(AIC(best.lm), 2),
    round(BIC(best.lm), 2),
    round(metrics_table["MAPE", best_name], 2),
    round(test_metrics["MAPE", best_name], 2),
    round(test_grade_summary[best_name, 1] * 100, 1)
  ),
  Trimmed = c(
    length(coef(trimmed.lm)) - 1,
    round(summary(trimmed.lm)$adj.r.squared, 4),
    round(AIC(trimmed.lm), 2),
    round(BIC(trimmed.lm), 2),
    round(mean(abs(trimmed.lm$residuals / train.clean$SalePrice)) * 100, 2),
    round(mean(abs((test.clean$SalePrice - trimmed.test) / test.clean$SalePrice)) * 100, 2),
    round(mean(assign_grade(abs(test.clean$SalePrice - trimmed.test) / test.clean$SalePrice) == "Grade 1: [0,0.10]") * 100, 1)
  )
)

kable(comparison_table, col.names = c("Metric", "Full (27 vars)", "Trimmed (17 vars)"),
      caption = "Full vs. Trimmed Model Comparison")

```

```{r}
cat("\n=== COLLINEARITY ASSESSMENT ===\n\n")

# Extract VIFs for the best model
vif_vals <- sort(vif(best.lm), decreasing = TRUE)
high_vif <- vif_vals[vif_vals > 4]

cat("Variables with VIF > 4:\n")
print(round(high_vif, 3))

cat("\nCorrelation between GarageCars and GarageArea:\n")
garage_cor <- cor(train.clean$GarageCars, train.clean$GarageArea)
cat("r =", round(garage_cor, 3), "\n\n")

cat("Assessment:\n")
cat("• GarageCars (VIF =", round(vif_vals["GarageCars"], 2), 
    ") and GarageArea (VIF =", round(vif_vals["GarageArea"], 2), 
    ") are correlated (r =", round(garage_cor, 3), ")\n")
cat("• Both VIFs are below the conventional threshold of 10\n")
cat("• Both variables are retained: GarageCars captures capacity (discrete)\n")
cat("  while GarageArea captures physical size (continuous) — they measure\n")
cat("  related but distinct aspects of garage value\n")
cat("• R² contribution: GarageCars =", 
    round(rsq_df$Pct_Change[rsq_df$Variable == "GarageCars"], 3), 
    "%, GarageArea =",
    round(rsq_df$Pct_Change[rsq_df$Variable == "GarageArea"], 3), "%\n")
cat("• If parsimony is prioritised, GarageArea is the weaker contributor\n")
cat("  and would be the candidate for removal\n")

```

```{r}
cat("\nNote on quality dummy VIFs:\n")
cat("• High VIFs for dummy variables from the same categorical variable are expected\n")
cat("  and not a concern — e.g. ExterQual_TA (VIF =", round(vif_vals["ExterQual_TA"], 2), 
    "), ExterQual_Gd (VIF =", round(vif_vals["ExterQual_Gd"], 2), ")\n")
cat("• Dummies from the same variable are structurally correlated by construction\n")
cat("• This is distinct from collinearity between independent predictors\n")
cat("  such as GarageCars and GarageArea, which warrants genuine concern\n")

```

```{r}
cat("\n=== FINAL MODEL SUMMARY ===\n\n")

cat("Selected Model:", best_name, "\n")
cat("Number of Predictors:", length(coef(best.lm)) - 1, "\n\n")

cat("Performance Metrics:\n")
final_summary <- data.frame(
  Metric = c("Adj R²", "AIC", "BIC", "Train MAPE (%)", "Test MAPE (%)", 
             "Test MSE", "Grade 1 Train (%)", "Grade 1 Test (%)", "Underwriting Quality"),
  Full = c(
    round(summary(best.lm)$adj.r.squared, 4),
    round(AIC(best.lm), 2),
    round(BIC(best.lm), 2),
    round(metrics_table["MAPE", best_name], 2),
    round(test_metrics["MAPE", best_name], 2),
    round(test_metrics["MSE", best_name], 2),
    round(underwriting["Train", best_name] * 100, 1),
    round(underwriting["Test", best_name] * 100, 1),
    ifelse(underwriting["Test", best_name] > 0.50, "YES", "NO")
  ),
  Trimmed = c(
    round(summary(trimmed.lm)$adj.r.squared, 4),
    round(AIC(trimmed.lm), 2),
    round(BIC(trimmed.lm), 2),
    round(mean(abs(trimmed.lm$residuals / train.clean$SalePrice)) * 100, 2),
    round(mean(abs((test.clean$SalePrice - trimmed.test) / test.clean$SalePrice)) * 100, 2),
    round(mean((test.clean$SalePrice - trimmed.test)^2), 2),
    round(mean(assign_grade(abs(trimmed.lm$residuals) / train.clean$SalePrice) == "Grade 1: [0,0.10]") * 100, 1),
    round(mean(assign_grade(abs(test.clean$SalePrice - trimmed.test) / test.clean$SalePrice) == "Grade 1: [0,0.10]") * 100, 1),
    ifelse(mean(assign_grade(abs(test.clean$SalePrice - trimmed.test) / test.clean$SalePrice) == "Grade 1: [0,0.10]") > 0.50, "YES", "NO")
  )
)
kable(final_summary, col.names = c("Metric", "Full (27 vars)", "Trimmed (17 vars)"), row.names = FALSE)

cat("\n\nKey Findings:\n")
cat("• TotalSqftCalc is the strongest predictor (6.21% R² contribution)\n")
cat("• Both models meet GSE underwriting quality standard (>50% Grade 1)\n")
cat("• Full model preferred: trimmed model loses 0.0086 Adj R², 4.9pp Grade 1, at cost of only 10 predictors\n")
cat("• Heteroscedasticity detected - consider log transformation for improvement\n")
cat("• Model generalizes well (test MAPE only 0.13% higher than training)\n")

```